---
title: "Tables"
description: |
  Tables for slides.
author:
  - name: Martin Frigaard
    url: https://www.martinfrigaard.io/
    affiliation: JOUR 301
    affiliation_url: https://mjfrigaard.github.io/FA22-JOUR301/
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
```

```{r IRAhandle_tweets_7_raw}
IRAhandle_tweets_7_raw <- readr::read_csv("https://raw.githubusercontent.com/fivethirtyeight/russian-troll-tweets/master/IRAhandle_tweets_7.csv")
glimpse(IRAhandle_tweets_7_raw)
```

| Variable      | Definition                                                                                                                                                                |
|---------------|---------------------------------------------------------------------------------------------------------------------------------------------------------------------------|
| ID            | 9-18 digit internal Twitter author account ID                                                                                                                             |
| Handle        | Twitter handle for author ("*JOHAAJOTS*", "*JOHANNBERGBERG*", "*JOHAN_NIGEL*")                                                                                            |
| Tweet Content | Tweet (i.e., "*#jobb Fake News Media Ignoring This! <https://t.co/kKF36W6Tfr> #job <https://t.co/BAJMuiSmke>*")                                                           |
| Region        | A region classification (i.e., "*United States*", "*Germany*", "*Azerbaijan*") determined by [Social Studio](https://help.salesforce.com/articleView?id=000199367&type=1) |
| Language      | The language of the tweet (i.e., "*English*", "*German*", "*Finnish*")                                                                                                    |
| Publish Date  | Date and time the original tweet was sent (i.e., "*8/5/2017 11:41*", "*8/5/2017 11:46*", "*8/5/2017 11:51*"                                                               |
