---
title: "Advanced Data Transformation and Manipulation"
comment: "*dealing with complex data structures*"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---

```{r , include=FALSE}
library(tidyverse)
library(skimr)
# chunk options ----
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      tidy = FALSE,
                      fig.path = "img/")
# base options ----
base::options(
  tibble.print_max = 25,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
```

# Objectives 

This document outlines some advanced data reshaping tasks with `tidyr`. Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) and [`tidyr`](https://tidyr.tidyverse.org/) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

# Materials 

View the slides for this section [here](https://mjfrigaard.github.io/csuc-data-journalism/slides.html).

## Load packages 

`tidyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

# Generating Sequences 

## `seq()` {.tabset}

We can use the `lubridate::ymd_hm()` function and `seq()` to create a sequence of dates or datetimes. 

### exercise 

Create a sequence of datetimes from midnight on `"2020-05-01"` until midnight on `"2020-05-02"`, separated by `"hour"`. 

Use `head()` to look at the first and last three values in `may_hours`.

```{r may_hours-ex, eval=FALSE}
may_hours <- seq(lubridate::ymd_hm("__________ 0:00"), 
    lubridate::ymd_hm("__________ 0:00"), 
    by = "____")
head(may_hours, _)
head(may_hours, _)
```

### solution

See below:

```{r may_hours-sol}
may_hours <- seq(lubridate::ymd_hm("2020-05-01 0:00"), 
    lubridate::ymd_hm("2020-05-02 0:00"), 
    by = "hour")
head(may_hours, 3)
head(may_hours, 3)
```

## `tidyr::full_seq()` {.tabset}

The `tidyr::full_seq()` works similar to `seq()`--it takes a vector (`x`), finds the `min` and `max` value, then generates the 'full sequence' of values, spaced out by the `period` argument.

Consider the vector `c(1, 3, 8, 10, 15)`.

```{r example-full_seq}
tidyr::full_seq(x = c(1, 3, 8, 10, 15), period = 1)
```

### exercise

Create a vector of two June dates: 

1. 8am on June 1, 2020.  
2. 5pm on June 1, 2020.

```{r jun_dates-ex, eval=FALSE}
jun_dates <- c(lubridate::ymd_hm("2020-__-__ __:__"), 
               lubridate::ymd_hm("2020-__-__ __:__"))
jun_dates
```

### solution

See below:

```{r jun_dates}
jun_dates <- c(lubridate::ymd_hm("2020-06-01 08:00"), 
               lubridate::ymd_hm("2020-06-01 17:00"))
jun_dates
```

### exercise 

Create a full sequence of June datetimes, starting at 8am June 1, 2020 and ending at 5pm June 1, 2020. Each value should be 15 min apart. 

```{r 15-min-ex, eval=FALSE}
head(
  full_seq(x = ________, 
           period = ___)
)
```

### solution

15 min x 60 seconds = 900 seconds. 

```{r 15-min-sol}
head(
  full_seq(x = jun_dates, 
           period = 900)
)
```


# Identify Missing Values 

We're going to use `expand_grid()` and `anti_join()` to find the missing values in the `VisitNAData` data frame. 

```{r load-VisitNAData, message=FALSE, warning=FALSE}
VisitNAData <- readr::read_csv(file = "https://bit.ly/36tnN55")
VisitNAData
```


## Creating List of Unique Values {.tabset}

Create a list of possible visit types (`"Consult"`, `"Lab"`, `"Clinic"`, `"Pharmacy"`) and store it in `visits`.

### exercise

```{r visits-ex, eval=FALSE}
______ <- c("______", "___", "_______", "__________")
```

### solution

See below:

```{r visits-sol}
visits <- c("Consult", "Lab", "Clinic", "Pharmacy")
```

## Expanding Data {.tabset}

Create a data frame with all possible combinations of `may_hours` and `visits`. Name the columns `date` and `visit_type`.

### exercise 

```{r FullDateVisit-ex, eval=FALSE}
FullDateVisit <- expand_grid(
  date = _______,
  visit_type = ________)
head(FullDateVisit)
```

### solution 

See below:

```{r FullDateVisit-sol}
FullDateVisit <- expand_grid(
  date = may_hours,
  visit_type = visits)
head(FullDateVisit)
```

## Anti-join {.tabset}

Check how many missing `visit_type`s there are with `dplyr::count()`

```{r VisitNAData-sol}
VisitNAData %>% 
  count(visit_type)
```

### exercise 

Join the `FullDateVisit` to `VisitNAData` with an anti-join, using `date` and `visit_type` as the id. 

```{r anti_join-ex, eval=FALSE}
anti_join(x = FullDateVisit, 
          y = ___________,
          by = c("date", "visit_type"))
```

### solution

These are the eight values that are missing from the `count()` output.

```{r anti_join-sol}
anti_join(x = FullDateVisit, 
          y = VisitNAData,
          by = c("date", "visit_type"))
```

# Complete Missing Values 

Let's return to our lab data from the previous exercises. Assume we have an data frame with an incomplete list of lab counts. 

```{r LabCounts-sol}
LabCounts <- tibble::tribble(
  ~lab, ~lab_count,
  "Albumin", 10L,
  "Potassium, Whole Blood", 8L,
  "% Hemoglobin A1c", 29L)
LabCounts
```

## Complete {.tabset}

One lab is missing from this list (`eGFR`), so we will make a vector with the known labs that are supposed to be the `LabCounts` data frame. 

### exercise 

```{r labs_list-ex, eval=FALSE}
labs_list <- c("Albumin", "Potassium, Whole Blood", 
               "% Hemoglobin A1c", "____")
labs_list
```

### solution

Now we have all five labs in `labs_list`.

```{r labs_list-sol}
labs_list <- c("Albumin", "Potassium, Whole Blood", 
               "% Hemoglobin A1c", "eGFR")
labs_list
```

### exercise 

Use `complete()` to fill in the missing `lab_count` for `eGFR`.

```{r complete-ex, eval=FALSE}
LabCounts %>%
  complete(
    lab = ________,
    fill = list(lab_count = __))
```

### solution

See below:

```{r complete-sol}
LabCounts %>%
  complete(
    lab = labs_list,
    fill = list(lab_count = NA))
```

### exercise 

Use `complete()` again, but use `list()` to make the missing `lab_count` values `0`.

```{r complete-2-ex, eval=FALSE}
LabCounts %>%
  complete(
    lab = ________,
    fill = list(lab_count = __))
```

### solution 

Note the `L` behind the `0`.

```{r complete-2-sol}
LabCounts %>%
  complete(
    lab = labs_list,
    fill = list(lab_count = 0L))
```

# Combine all three!

We're going to combine the use of `pivot_longer()`, `complete()`, and `full_seq()`. 

## `difftime` {.tabset}

Assume we have a dataset (`Infected`) that contains information on 20 patients (`patient`), the date they became `infected_date`, and the date they infection cleared (`recovered_date`). 

```{r Infected, message=FALSE, warning=FALSE}
Infected <- read_csv("https://bit.ly/33zp4Fw")
Infected
```

### exercise 

What if we wanted to know the number of days between `infection_start` and `infection_end`? One way to do this is with `difftime()`. It takes two datetime values, and the units the differences should be in (`"days"`)

Calculate the number of days between `infection_start` and `infection_end` in days.

```{r difftime-ex, eval=FALSE}

Infected %>% 
  mutate(infected_days = difftime(time1 = _____________, 
                                  time2 = _____________,
                                  units = "days"))
```

### solution

See below:

```{r difftime-sol}
Infected %>% 
  mutate(infected_days = difftime(time1 = infection_end, 
                                  time2 = infection_start,
                                  units = "days"))
```

## `pivot_longer` {.tabset}

The previous example gave us the total number of days for each patient's infection, but what if we wanted to calculate a running total of infected days for each patient? We would have to reshape the `Infected` dataset into a longer format. 



### exercise 

Pivot the two date columns in the `Infected` dataset, assign the names to `"infection_status"`, and the values to `"date".` 

```{r pivot_longer-ex, eval=FALSE}
Infected %>% 
  pivot_longer(-patient, 
  names_to = "_____________", 
  values_to = "____")
```

### solution

See below:

```{r pivot_longer-sol}
Infected %>% 
  pivot_longer(-patient, 
  names_to = "infection_status", 
  values_to = "date")
```

## `complete` {.tabset}

Now that we have `date` and `infection_status` in a long format, we can start calculating the number of days between `infection_start` and `infection_end`.


### exercise

Group the data by `patient`, then `complete()` the dataset: 

- assign `patient` to `patient`  
- set `date` using the `full_seq()` function to fill in the missing days between the `infection_start` and `infection_end` dates   
- use the `fill` argument to replace the missing values in `infection_status` as `"infected"`.

```{r complete-infection-date-ex, eval=FALSE}
Infected %>% 
  pivot_longer(-patient, 
  names_to = "infection_status", 
  values_to = "date") %>% 
  group_by(_________) %>% 
  complete(patient = _______, 
           date = full_seq(____, period = 1),
           fill = list(infection_status = "___________"))
```

### solution

See below:

```{r complete-infection-date-sol}
Infected %>% 
  pivot_longer(-patient, 
  names_to = "infection_status", 
  values_to = "date") %>% 
  group_by(patient) %>% 
  complete(patient = patient, 
           date = full_seq(date, period = 1),
           fill = list(infection_status = "infected")) -> InfectedLong
InfectedLong
```

## `lag()` {.tabset}

In order to calculate the running total of infected days, we're going to combine `difftime()`, `if_else()`, and `cumsum()`.

### exercise 

We're going to start by using `difftime()` and `lag()` to calculate the difference between `date` and `lag(x = date, n = 1)`. 

Instead of using `"days"`, we're going to pass the output to `as.numeric()` so it's converted to a `double`.

```{r difftime-2-ex, eval=FALSE}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = ____, 
                                    time2 = lag(x = ____, 
                                                n = 1))))
```

### soltution 

See below:

```{r difftime-2-sol}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))))
```

## `if_else()` {.tabset}

### exercise 

Create `day_diff` and use `dplyr::if_else()` to replace the `NA` in `day_diff0`. 

`dplyr::if_else()` takes a logical condition, the desired result if it is `true`, and the desired result if it is `false`


```{r if_else-ex, eval=FALSE}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
    # replace missing day_diff0 with 0, all else get day_diff0
    day_diff = dplyr::if_else(condition = is.na(________), 
                      true = _, 
                      false = ________))
```

### solution

See below:

```{r if_else-sol}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
    day_diff = dplyr::if_else(condition = is.na(day_diff0), 
                      true = 0, 
                      false = day_diff0))
```

No we have `0`'s instead of `NA`s.

## `cumsum()` {.tabset}

### exercise 

Finally we have a numerical variable we can add up to get the number of days infected at each day in `date`. Here we will use the `cumsum()` function (cumulative sum) to get the running total number of days infected per `patient` (recall the dataset is still grouped by `patient`).

Create `total_inf_days` by getting the `cumsum()` of a numerically formatted `day_diff` column.

```{r cumsum-ex, eval=FALSE}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
     day_diff = dplyr::if_else(condition = is.na(day_diff0), 
                      true = 0, 
                      false = day_diff0), 
    
     _____________ = cumsum(as.numeric(________)))
```

### solution

See below

```{r cumsum-sol}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
     day_diff = dplyr::if_else(condition = is.na(day_diff0), 
                      true = 0, 
                      false = day_diff0), 
    
   total_inf_days = cumsum(as.numeric(day_diff)))
```


## Clean Up {.tabset}

These last few steps are for cleaning up our `InfectedLong` by removing the underscore (`_`) from `infection_status`, `ungrouping()` the data, and dropping the intermediate `day_diff0` and `day_diff` columns.

### exercise

- use `stringr::str_replace_all()` to remove the underscores by giving `infection_status` to the `string` argument, `"_"` to the `pattern` argument, and `" "` to the `replacement` argument  

- `ungroup()` the data  

- use `select()` to remove the 

```{r ungroup-select-ex, eval=FALSE}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
     day_diff = dplyr::if_else(condition = is.na(day_diff0), 
                      true = 0, 
                      false = day_diff0), 
    
   total_inf_days = cumsum(as.numeric(day_diff)),
   
   infection_status = str_replace_all(string = __________________, 
                                      pattern = ___, 
                                      replacement = ___)) %>% 
  ________() %>% 
  select(-c(_________, _________))
```

### solution

See below.

```{r ungroup-select-sol}
InfectedLong %>% 
  mutate(
    day_diff0 = as.numeric(difftime(time1 = date, 
                                   time2 = lag(x = date, 
                                               n = 1))),
     day_diff = dplyr::if_else(condition = is.na(day_diff0), 
                      true = 0, 
                      false = day_diff0), 
    
   total_inf_days = cumsum(as.numeric(day_diff)),
   
   infection_status = str_replace_all(string = infection_status, 
                                      pattern = "_", 
                                      replacement = " ")) %>% 
  ungroup() %>% 
  # remove intermediate columns
  select(-c(day_diff0, day_diff)) -> InfectedLong

InfectedLong
```

### Plotting Counts

Now we can count the number of times each `date` occurs and plot if over `date`.

```{r plotting-counts}
InfectedLong %>% 
  count(date, name = "days_inf") %>% 
  ggplot(aes(x = date, y = days_inf)) +
  geom_line() + 
  labs(x = "Date", y = "Days Infected", 
       title = "Distribution of Infected Days Over Time")
```

