---
title: "Reshaping Data in R"
description: |
  "*separate, separate rows, and unite data with `tidyr`*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 3
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
library(fivethirtyeight)
library(fivethirtyeightdata)
# create data folder
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 8.5,
  fig.height = 6,
  fig.retina = 2
  )
fs::dir_create("data")
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/reshape-join/')
# tibble printing
options(
  pillar.print_max = 10, 
  pillar.print_min = 10)
```

# Objectives 

This document outlines some common data reshaping tasks with the  `tidyr` and `dplyr` packages. **If you've ever Googled 'how to perform [x] in R', you've probably discovered there are multiple ways to accomplish [x].** The methods I present here are not exhaustive, and I've tried to balance the trade-off between 1) solutions/functions that solve problems with the fewest number of keystrokes, and 2) code that's readable, explicit, and easy to follow.

## Load packages 

`tidyr` and `dplyr` are part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(fivethirtyeight)
```

## Namespacing

I'll be using the namespace syntax (`package::function()`) just to make it explicitly clear which function belongs to which package, but this isn't necessary if you've run the code block above. 


## Tidy data

There are three principles for tidy data: 

1. Columns hold *variables*    
2. Rows hold *observations*    
3. Cells (the intersection of rows and columns) hold *values*  

We're going to be using small samples of three datasets from the [`fivethirtyeight` package](https://fivethirtyeight-r.netlify.app/), and the `BloodPressures` dataset. 


### `EndorseSample`

There are multiple ways data can be 'less-than tidy.' We'll start with a common example, which is two variables (and their values) stored in a single column.  

Below is a subset of the `endorsements_2020` dataset named `EndorseSample`. Read more about the `endorsements_2020` dataset by typing `??endorsements_2020` into the console. 

```{r EndorseSample, message=FALSE, warning=FALSE}
EndorseSample <- readr::read_csv("https://bit.ly/3S7gYuh")
dplyr::glimpse(EndorseSample)
```

Notice the `location` variable has two values (`state` and `city`) stored in the same column. 

### `FandangoSample`

In another example of un-tidy data, we'll look at the dataset below which comes from the `fandango` dataset. 

```{r FandangoSample, message=FALSE, warning=FALSE}
FandangoSample <- readr::read_csv(file = "https://bit.ly/3Tg1zt6")
dplyr::glimpse(FandangoSample)
```

This dataset has two *similar* scores combined into a single column `rottentomatoes_scores`

+ `user_perc`: The Rotten Tomatoes user score for the film (expressed as a percent)

+ `user_norm`: The Rotten Tomatoes user score for the film , normalized to a 0 to 5 point system (expressed as 'score `out of 5`')

### `DailyShowSample`

Below is a sample from the `daily_show_guests` dataset. The guests names are separated into `first_name` and `last_name` columns. 

```{r DailyShowSample, message=FALSE, warning=FALSE}
DailyShowSample <- readr::read_csv("https://bit.ly/3yR2PuA")
dplyr::glimpse(DailyShowSample)
```

### `BloodPressures`

The `BloodPressures` dataset has names, blood pressures, and units for five people. These data contain 1) multiple measurements in one column, and 2) multiple units in one column.  

```{r BloodPressures, eval=TRUE, message=FALSE, warning=FALSE}
BloodPressures <- readr::read_csv("https://bit.ly/3gjZ0rl")
dplyr::glimpse(BloodPressures)
```

***

# Reshaping Columns and Rows 

In the next few exercises, we'll go over how (and *when*) to use `separate()`, `separate_rows()`, and `unite()`. These functions are typically used in a data wrangling 'pipeline', so many of these examples have been reduced to focus on on the function arguments. 

## `separate()` 

This allows us to split columns apart into multiple variables. 

Take another look at `EndorseSample` below:

```{r glimpse-EndorseSample, eval=TRUE}
dplyr::glimpse(EndorseSample)
```

### Arguments

The arguments for `separate_rows()` are below:

```{r separate_args, eval=FALSE}
tidyr::separate(data = , col = , into = , sep = , remove = TRUE, 
         convert = FALSE, extra = "warn", fill = "warn", ... = )
```

+ The `data` is whatever table we're looking to wrangle  

+ `col` are the columns that need separating, and `into` should contain the names of the new columns we want to create with the separated values  

+ The `sep` argument is the *separator between columns.*, and it defaults to `"[^[:alnum:].]+"` (which will match on just about any number or punctuation)  

+ `remove` controls what to do with the column after we've separated the values (defaults to `TRUE`, but I initially like to set to `FALSE` to review)  

+ `extra` controls what will happen with extra pieces (defaults to `"warn"`, but other options are `"drop"` or `"merge"`). See `??separate()` for more info. 

+ `fill` is similar to `extra`, but controls what will happen when there aren't enough values in the column we're separating. Defaults to `"warn"`, but other options are `"right"` or `"left"`). See `??separate()` for more info. 

The code below separates `location` into `state` and `city`.

```{r separate-sol}
EndorseSample %>% 
    tidyr::separate(col = location, 
             into = c("state", "city"), 
             sep = ", ", 
             remove = FALSE) %>% 
    dplyr::select(location, state, city)
```

The `sep` argument is the character we want to separate the values on (in our case, it's a comma with a space: `", "`), and we add the `remove = FALSE` so the `location` variable stays in the dataset. 

We can also see by adding `select()` with 1) the original column, and 2) the two columns we created to the end of the pipeline, we're able to view the result before assigning it to an output. 

***

## `separate_rows()` 

`separate_rows()` is similar to `separate()`, but drops the values down the rows. This is commonly used when we have more than one *similar* value in each column, or a ratio/fraction.

### Arguments 

The arguments for `separate_rows()` are below: 

```{r separate_rows_args, eval=FALSE}
tidyr::separate_rows(data = , ... = , sep = , convert = FALSE)
```

+ The `data` is whatever table we're looking to wrangle 

+ The `...` is a placeholder for the columns we want to separate across the rows 

+ The `sep` argument is similar to `separate()` (*separator delimiting collapsed values*), and it defaults to the same regular expression `"[^[:alnum:].]+"`  

+ `convert` controls whether to convert the columns to a specific type (i.e., `numeric`, `integer`, etc.). Defaults to `FALSE`. 

Take a look at `BloodPressures` below: 

```{r head-BloodPressures}
utils::head(BloodPressures)
```

The code below separates the `blood_pressure` into two values, but places the second value on the following row:  

```{r separate_rows_ex_01}
BloodPressures |> 
  tidyr::separate_rows(blood_pressure) |> 
  dplyr::select(blood_pressure, units)
```

### The `sep` argument

Notice that we omitted the `sep` argument and it still separated the values on the correct character? That's because the default `sep` pattern (`"[^[:alnum:].]+"`) will match the correct character (`"/"`). 

### Multiple columns

One issue with the output above is that the `units` are still in the previous format (`systolic (mm Hg) / diastolic (mm Hg)`). Fortunately, we can just add a second variable to the first argument in `separate_rows()` (as a vector in `c()`). 

```{r separate_rows_ex_02, error=TRUE}
BloodPressures |> 
  tidyr::separate_rows(
    c(blood_pressure, units)) 
```

We get an error because we have to provide the `sep` argument (and it's a good habit to include it, just to be more explicit).

```{r separate_rows_ex_03}
BloodPressures |>
  tidyr::separate_rows(c(blood_pressure, units),
    # include sep here
    sep = " / ") |> # don't forget the spaces!
  dplyr::select(blood_pressure, units)
```

Now we have both `blood_pressure` and `units` in their own columns. We would probably want to `rename` these variables something more precise (with names that will make it easier to locate variables in future pipelines (i.e., using `starts_with("bp")`). 

```{r separate_rows_ex_04}
BloodPressures |>
  tidyr::separate_rows(c(blood_pressure, units), sep = " / ") |>
  dplyr::rename(
    bp = blood_pressure, # new column names! 
    bp_units = units) |>
  dplyr::select(starts_with("bp"))
```

## `unite()` 

Just like `separate`/`separate_rows` allows us to split up columns, `unite()` lets us stick them back together! Take a look at `DailyShowSample`.

```{r review-DailyShowSample}
dplyr::glimpse(DailyShowSample)
```

We can see the guests have had their names in separate columns (`first_name` and `last_name`).

The code below unites `first_name` and `last_name` into `guest_name` 

```{r unite_ex_01}
DailyShowSample |>  
    unite(col = "guest_name", 
          first_name:last_name, 
          sep = " ", 
          remove = FALSE) |> 
  dplyr::select(dplyr::contains("name"))
```

We chose to keep the previous columns by adding `remove = FALSE`

## Verifying calculated values 

When encountering a new data source, it's a good practice to check any calculated values. It's not necessary to verify *every* column and row, but if a column have been derived from columns containing measurements, you'll want to verify these before building any visualizations or tables. 

For example, if a dataset dplyr::contains measurements for `height` and `weight`, and a calculated `BMI` column, we'd want to verify the BMI using whatever methods were reported with the data. 

We'll walk through an example with the `FandangoSample` below:

```{r re-glimpse-FandangoSample}
dplyr::glimpse(FandangoSample)
```

We're going to separate the two scores in `rottentomatoes_scores`, then verify these scores by re-creating the columns. 

- **Use `tidyr::separate()` to separated `rottentomatoes_scores` into `user_perc` and `user_norm`**   
  - *Add `remove = FALSE` to keep `rottentomatoes_scores` so we can check our work.*  

```{r verify_ex_01}
FandangoSample |> 
  # separate scores
  tidyr::separate(col = rottentomatoes_scores, 
    into = c("user_perc", "user_norm"), 
    sep = "/", remove = FALSE) |> 
  dplyr::select(dplyr::contains("user"))
```

- **Now we want to use `dplyr::mutate()` and `stringr::str_remove_all()` to remove the text (`" out of 5"`)**
  - *the `string` argument is the column we want to alter, the `pattern` argument is the text we want to remove.*
  
```{r verify_ex_02}
FandangoSample |>
  tidyr::separate(
  # separate scores
    col = rottentomatoes_scores,
    into = c("user_perc", "user_norm"),
    sep = "/",
    remove = FALSE
  ) |>
  dplyr::mutate(
    # remove text
    user_norm = stringr::str_remove_all(
      string = user_norm, pattern = " out of 5")
    ) |>
  dplyr::select(dplyr::contains("user"))
```

- **Now that we've removed the text, we can format the `user_norm` as a `double` column**    
  - *recall that `double` vectors can have decimals*   
  
```{r verify_ex_03}
FandangoSample |> 
  tidyr::separate(col = rottentomatoes_scores, 
    into = c("user_perc", "user_norm"), 
    sep = "/", remove = FALSE) |> 
  dplyr::mutate(
    # remove text
    user_norm = stringr::str_remove_all(
      user_norm, " out of 5"), 
    # convert to double
    user_norm = as.double(user_norm)
    ) |> 
  dplyr::select(dplyr::contains("user"))
```

- **To verify the scores calculated in `user_perc`, we're going to create a new column, `user_perc_check`**  
  - *the calculation is fairly straightforward--it's in the text we removed!*  
  
```{r verify_ex_04}
FandangoSample |> 
  tidyr::separate(col = rottentomatoes_scores, 
    into = c("user_perc", "user_norm"), 
    sep = "/", remove = FALSE) |> 
 dplyr::mutate(
   # remove text
    user_norm = stringr::str_remove_all(
      string = user_norm, pattern = " out of 5"), 
   # convert to double
    user_norm = as.double(user_norm), 
   # check calculation
    user_perc_check = user_norm/5.00  # 'out of 5'
   ) |>
  dplyr::select(dplyr::contains("user"))
```

Like the adage goes, [**Trust, but verify**](https://en.wikipedia.org/wiki/Trust,_but_verify). If the calculations are complicated or require further explanation--reach out to the original party. As a general rule, if someone can't explain to you how they arrived at a particular value, you shouldn't trust it. 

`unite()` also comes in handy when we want to create graph labels for data values. If we wanted to create a column similar to  `rottentomatoes_scores`, but with our verified column, we could do the following: 

```{r}
FandangoSample |> 
  tidyr::separate(col = rottentomatoes_scores, 
    into = c("user_perc", "user_norm"), 
    sep = "/", remove = FALSE) |> 
 dplyr::mutate(
   # remove text
    user_norm = stringr::str_remove_all(
      string = user_norm, pattern = " out of 5"), 
   # convert to double
    user_norm = as.double(user_norm), 
   # check calculation
    user_perc_check = user_norm/5.00  # 'out of 5'
   ) |>
  # add film and year
  tidyr::unite(col = "label", 
    c(film, year), sep = ", ", 
    remove = FALSE) |> 
  # add user norm values 
  tidyr::unite(col = "label", 
    c(label, user_norm), sep = " (", 
    remove = FALSE) |> 
  # add verified user percent
  tidyr::unite(col = "label", 
    c(label, user_perc_check), 
    sep = " out of 5) or ", 
    remove = FALSE) |> 
  dplyr::select(label)
```


***

