---
title: "Joining Data in R"
description: |
  "*Using `dplyr` to join data tables*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 3
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(inspectdf)
library(starwarsdb)
library(dm)
# create data folder
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5.5,
  fig.retina = 2
  )
fs::dir_create("data")
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/reshape-join/')
# tibble printing
options(
  pillar.print_max = 10, 
  pillar.print_min = 10)
```

# Objectives 

This document outlines some common data joining tasks with the `dplyr` package. **If you've ever Googled 'how to perform [x] in R', you've probably discovered there are multiple ways to accomplish [x].** The methods I present here are not exhaustive, and I've tried to balance the trade-off between 1) solutions/functions that solve problems with the fewest number of keystrokes, and 2) code that's readable, explicit, and easy to follow.


## Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. We'll also be installing the `inspectdf`, `starwarsdb`, and `dm` packages. 

```{r packages, eval=FALSE}
# install.packages(c("tidyverse", "inspectdf", "starwarsdb", "dm"))
library(tidyverse)
library(inspectdf)
library(starwarsdb)
library(dm)
```

## Namespacing

I'll be using the namespace syntax (`package::function()`) just to make it explicitly clear which function belongs to which package, but this isn't necessary if you've run the code block above. 

## Inspecting data tables

Whenever I'm performing a series of joins, I like to use the [`inspectdf` package](https://alastairrushworth.github.io/inspectdf/). This package has a series of functions for viewing all the columns in a table by various types: 

+ **`inspectdf::inspect_cat()`**: display categorical variables (missing values are displayed by color)  

+ **`inspectdf::inspect_cor()`**: display the correlation between numerical variables  

+ **`inspectdf::inspect_imb()`**: display imbalances between variables    

+ **`inspectdf::inspect_mem()`**: display the memory usage by column  

+ **`inspectdf::inspect_na()`**: display the rate of missingness by column  

+ **`inspectdf::inspect_num()`**: display the distribution of the numeric columns.

+ **`inspectdf::inspect_types()`**: display the column types.  

+ **`inspectdf::show_plot()`**: show the plot from the `inspect_*` function  
  - we'll also add the **`text_labels = TRUE`**, **`col_palette = 1`** (for the  colorblind friendly palette).


There are three principles for tidy data: 

1. Columns hold *variables*    
2. Rows hold *observations*    
3. Cells (the intersection of rows and columns) hold *values* 


# Joins

Joins give us the ability to combine multiple datasets on a common column. We're going to be using the [`starwarsdb` package](https://github.com/gadenbuie/starwarsdb), which dplyr::contains data from the [Star Wars API](https://swapi.dev/). 

## A Data Model 

The code below creates and displays the data model for the tables in this package 

```{r sw_data_model, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
sw_data_model <- dm::dm_draw(dm = starwarsdb::starwars_dm(),
  graph_name = "sw_data_model",
  view_type = "all")
sw_data_model
```

As you can see from the graph above, there are nine tables in the `starwarsdb`. The tables connect to each other through a series of common columns (called `keys`), and these allow us to perform multiple kinds of joins. 

## Create sw_films

Load the `starwarsdb::films` dataset, removing the `opening_crawl` column, and filter to only episodes 4-6. Store these data in `sw_flms`

```{r sw_flms}
sw_flms <- starwarsdb::films |> 
  dplyr::select(-c(opening_crawl, director, producer)) |> 
  dplyr::filter(episode_id >= 4 & episode_id <= 6)
sw_flms
```


## `left_join`

***A `left_join` keeps all of `x`, and joins it to all matching rows from dataset `y`***

We want to left-join `sw_flms` to `starwarsdb::films_vehicles`: 

```{r films_vehicles}
starwarsdb::films_vehicles
```

We can do this on `"title"`, then we can re-organize the columns using `dplyr::select()` to place the `vehicle` column before `dplyr::everything()` else. 

```{r left_join_ex_01}
dplyr::left_join(x = sw_flms, 
          y = starwarsdb::films_vehicles, 
          by = "title")  |>   
  dplyr::select(vehicle, 
    dplyr::everything()) 
```


Assign to `sw_flms_veh`

```{r left_join_ex_02}
dplyr::left_join(x = sw_flms, 
          y = starwarsdb::films_vehicles, 
          by = "title")  |>   
  dplyr::select(vehicle, 
    dplyr::everything()) -> sw_flms_veh
```

Let's take a look at the sw_flms_veh table:

```{r}
sw_flms_veh |> dplyr::glimpse()
```

We can see this table has 4 columns and 47 rows. We'll use the `inspect_types()` function below to view the breakdown of the columns by `type`: 

```{r sw_flms_veh-inspect_types}
sw_flms_veh |> 
  inspectdf::inspect_types() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1)
```

We can see most of these columns are character. Below we'll display the categorical variables using `inspectdf::inspect_cat()`:

```{r sw_flms_veh-inspect_cat}
sw_flms_veh |> 
  inspectdf::inspect_cat() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1)
```

```{r}
sw_flms_veh |> 
  inspectdf::inspect_num() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1) 
```


## `inner_join`

***The `inner_join()` joins dataset `x` and `y`, and keeps only matching rows from both.***

The code below uses an `inner_join` to join `sw_films_veh` to `starwarsdb::pilots` on `"vehicle"`, and assigns it to `sw_flms_veh_plt`

```{r sw_flms_veh_plt-sol}
dplyr::inner_join(x = sw_flms_veh, 
           y = starwarsdb::pilots, 
           by = "vehicle") -> sw_flms_veh_plt
sw_flms_veh_plt |> dplyr::glimpse()
```

```{r sw_flms_veh_plt-inspect_types}
sw_flms_veh_plt |> 
  inspectdf::inspect_types() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1)
```

```{r sw_flms_veh_plt-inspect_cat}
sw_flms_veh_plt |> 
  inspectdf::inspect_cat() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1)
```

## `right_join`

***The `right_join()` includes all the rows in dataset `y`, and joins it the matching rows in dataset `x`***

The code below uses a `right_join()` to join `sw_flms_veh_plt` to `starwarsdb::films_planets` on `"title"`, keeping all the rows in `starwarsdb::films_planets`. 

Then we use `select()` to rename `planet` to `flm_planet` and assign this to `sw_flms_veh_plt_plnt` 

```{r right_join-sol}
dplyr::right_join(
  x = sw_flms_veh_plt,
  y = starwarsdb::films_planets,
  by = "title") %>%
  dplyr::select(
    vehicle:release_date,
    film_planet = planet,
    pilot) -> sw_flms_veh_plt_plnt
sw_flms_veh_plt_plnt |> dplyr::glimpse()
```

```{r sw_flms_veh_plt_plnt-inspect_types}
sw_flms_veh_plt_plnt |>
  inspectdf::inspect_types() |>
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1)
```

```{r sw_flms_veh_plt_plnt-inspect_cat}
sw_flms_veh_plt_plnt |> 
  inspectdf::inspect_cat() |> 
  inspectdf::show_plot(text_labels = TRUE, 
    col_palette = 1)
```

## `anti_join`

***The `anti_join()` keeps all rows in dataset `x` that **do not** have a match in dataset `y`.***

Note the last join created empty values for `vehicle`, `episode_id`, `release_date`, and `pilot`:

```{r filter-is.na}
sw_flms_veh_plt_plnt |> 
  dplyr::filter(is.na(episode_id)) |> 
  inspectdf::inspect_na() |> 
  inspectdf::show_plot(text_labels = TRUE, 
    col_palette = 1) +
  coord_flip()
```

The code below uses the `anti_join()` to join `sw_flms_veh_plt_plnt` to `sw_films` and create a dataset with only `title` and `film_planet`:

```{r anti_join-inspect_types}
dplyr::anti_join(x = sw_flms_veh_plt_plnt, 
            y = sw_flms, 
            by = c("title", "episode_id", "release_date")) |> 
  dplyr::select(title, film_planet) |> 
  inspectdf::inspect_types() |>
  inspectdf::show_plot(text_labels = TRUE, col_palette = 1)
```


```{r anti_join-inspect_cat}
dplyr::anti_join(x = sw_flms_veh_plt_plnt, 
            y = sw_flms, 
            by = c("title", "episode_id", "release_date")) |> 
  inspectdf::inspect_na() |> 
  inspectdf::show_plot(
    text_labels = TRUE, col_palette = 1) + 
  ggplot2::coord_flip()
```

These should only contain the `title`s **not** in `sw_films`.