---
title: "Scraping Wikipedia Tables"
description: |
  "*Advanced wrangling example*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2

always_allow_html: true

---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
# chunk options ----
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 8.5,
  fig.height = 6,
  fig.retina = 2
  )
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/scrape-wikipedia/')
fs::dir_create("data")
# base options ----
base::options(
  tibble.print_max = 25,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
```

# Objectives

This document outlines some advanced data wrangling with [`dplyr`](https://dplyr.tidyverse.org/index.html), [`tidyr`](https://tidyr.tidyverse.org/index.html), and [`stringr`](https://stringr.tidyverse.org/index.html). Read more in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science.

## Load packages

These are all part of the `tidyverse`, so we install and load this meta-package below.

```{r packages}
# install.packages("tidyverse")
library(tidyverse)
```

# Pharmaceutical Companies

In the slides we had well-formatted dataset. In these exercises, we're going to import some data from the wild (Wikipedia), because most data aren't ready to visualize and model when we first get them.

## Import the data 

I Googled "largest pharmaceutical companies" and found [this wikipedia page](https://en.wikipedia.org/wiki/List_of_largest_biotechnology_and_pharmaceutical_companies). 

These packages will help us scrape the data in the table and manipulate it in R. 

```{r scrape-packages, message=FALSE, warning=FALSE}
library(xml2)
library(rvest)
```


### Read HTML

The first function I'll use comes from the `xml2` [package](https://blog.rstudio.com/2015/04/21/xml2/). `xml2::read_html()` loads the html from the wikipedia page into an R object I call `wiki_html`.

```{r wiki_html}
wiki_html <- xml2::read_html("https://en.wikipedia.org/wiki/List_of_largest_biotechnology_and_pharmaceutical_companies")
```

We should check the structure of new objects, so we know what we're working with.

```{r structure_wiki_html}
wiki_html %>% str()
```

I can see this is a list of two objects (a `node` and a `doc`).

### HTML Nodes (css)

I want the html `node`, so I will use the `html_nodes()` function from the `rvest` [package](https://github.com/hadley/rvest). 

The `css` argument is set to `"table"`. Once again I check the structure of the output object.

```{r wiki_html_tables}
wiki_html_tables <- wiki_html %>% rvest::html_nodes(css = "table")
wiki_html_tables %>% str()
```

This is a list of three lists, each of them `xml_node`s. 

In order to find the relevant tables in the `wiki_html_tables` object, I need to be able to search on something. Fortunately, the `base::grep()` function can be used in combination with sub-setting to extract the `relevant_tables` from `wiki_html_tables`.

Get the relevant tables from the `xml_nodeset` in `wiki_html_tables`.

```{r relevant_tables}
relevant_tables <- wiki_html_tables[base::grep("rank", wiki_html_tables)]
relevant_tables %>% str()
```

This returned yet another list of lists! (another set of `node` and `doc`). 

### Extract HTML table

Now I need to use `rvest::html_table()` function (combined with some bracket sub-setting) to explore this object and learn about it's contents (read more about how this works [here](https://dcl-wrangle.stanford.edu/rvest.html#scrape-data-with-rvest)). 

I will start with position `[[1]]` and set `fill = TRUE`. 

```{r test-html_table}
rvest::html_table(relevant_tables[[1]], 
                  fill = TRUE) %>% utils::head()
```

This looks like the table I want! Let's assign it to a data frame and get wrangling!

```{r TopPharmCompRaw}
TopPharmCompRaw <- rvest::html_table(relevant_tables[[1]], 
                  fill = TRUE)
TopPharmCompRaw
```

```{r export-TopPharmCompRaw}
fs::dir_create("data/wikipedia-data/")
wikipedia_top_pharm_export_file <- paste0("data/wikipedia-data/", lubridate::today(), "-TopPharmCompRaw.csv")
readr::write_csv(x = TopPharmCompRaw, file = wikipedia_top_pharm_export_file)
fs::dir_tree("data/wikipedia-data/")
```

```{r import-TopPharmCompRaw, include=FALSE, eval=TRUE}
TopPharmCompRaw <- readr::read_csv("data/wikipedia-data/2020-11-24-TopPharmCompRaw.csv")
```

## Outline before you start 

We have a raw dataset now, and we should make a 'to-do' list of what we want the data to look like. I start with data tidying, then move into changes for the individual variables. 

1. Restructure the data (as needed)  
2. Create column names  
3. Wrangle/format variables 

# Wrangle Data 

These data are in a wide format, with each `market_cap_in_` spread across columns. 

```{r view-TopPharmCompRaw}
TopPharmCompRaw
```

## Pivot data 

Ideally, we want two variables, `year` and `market_cap_us_bil`. 

### `pivot_longer`

We can re-shape `TopPharmComp` with `tidyr::pivot_longer()`, but first we need all the variables in the *same* format. We can do this with `mutate_if()`.

```{r check-pivot_longer}
TopPharmCompRaw %>% 
  mutate_if(is.numeric, as.character) %>% 
  tidyr::pivot_longer(cols = starts_with("Market Cap in"), 
                      names_to = "year", 
                      values_to = "market_cap_us_bil")
```

This looks correct--let's assign it to `TidyTopPharmComp` (because it's a new data structure).

```{r assign-TidyTopPharmComp}
TidyTopPharmComp <- TopPharmCompRaw %>% 
  mutate_if(is.numeric, as.character) %>% 
  tidyr::pivot_longer(cols = starts_with("Market Cap in"), 
                      names_to = "year", 
                      values_to = "market_cap_us_bil")
```

## Format Column Names 

Follow along on each tab for the steps to wrangle the column names:

1. Each variable in it's own column:  
   - i.e. type of company in it's own column (`B` – Biotechnology company, `P` – Pharmaceutical company)
   - stock exchange identifier in it's own column (i.e. `NYSE`, `NASDAQ`, etc.) 
2. Properly formatted values (numeric, factor, etc.)   
3. Missing variables formatted correctly (replace `-` with `NA`)

### Names

We can use `janitor::clean_names()`, but first we should do some string manipulation to remove the extra characters from the column names. 

```{r cleaned_names}
TidyTopPharmComp %>% 
    # get a vector of 'dirty' names
    names() %>% 
    # remove citations in brackets [ ]
    stringr::str_remove_all(string = ., pattern = "\\[[0-9]\\]") %>%
    # replace (USD billions)
    stringr::str_replace_all(string = ., pattern = "\\s*\\([^\\)]+\\)", 
                             replacement = "_us_bil") %>% 
    # remove bracket from Rank
    stringr::str_remove_all(string = ., pattern = "\\[|\\]") %>% 
    # remove alpha numeric from end of Rank
    stringr::str_remove_all(string = ., pattern = "N 1$") %>% 
    # make all lowercase
    stringr::str_to_lower() -> cleaned_names
cleaned_names
```

Now we can apply the new names (`cleaned_names`) to `TidyTopPharmComp` and use `janitor::clean_names()` to standardize them. 

```{r clean_names}
TidyTopPharmComp %>% 
    set_names(nm = cleaned_names) %>% 
    janitor::clean_names()
```

### `set_names()` & `clean_names()`

If this looks acceptable, we can assign to `TopPharmComp`.

```{r TopPharmComp}
TopPharmComp <- TidyTopPharmComp %>% 
    set_names(nm = cleaned_names) %>% 
    janitor::clean_names()
```



## Define Variables 

We want to split up the following variables into their own columns: 

1. Year as a four-number digit `year`
2. `company_type` = `[P]` for `Pharmaceutical` or `[B]` for `Biotechnology`  
3. `stock_exch` = `NSYE` (New York Stock Exchange), `NASDAQ` (National Association of Securities Dealers Automated Quotations), `FWB` (Frankfurt Stock Exchange), `TYO` (Australian Securities Exchange), `TSX` (Toronto Stock Exchange), and `SIX` (Swiss Exchange).
4. `stock_id` = acronym for each company on `stock_exch`  
5. `largest_market_cap_date` = date from `largest_market_cap_us_bil`  
6. `company_name` a variable containing *only* the company name 
7. a `ranking` variable with a numerically coded rank

## `year` 

We can also wrangle the `market_cap_year` variable so it only contains the four-number year. 

### Remove characters 

Here we use the `\\D` regular expression to remove all non-numbers. 

```{r year}
TopPharmComp %>% 
  # remove characters
  dplyr::mutate(year = str_remove_all(string = year, pattern = "\\D"),
                year = str_sub(string = year, start = 1L, end = 4L),
                # make numeric
                year = as.integer(year))
```

### `year` 

Let's assign `year` to `TopPharmComp`

```{r assign-year}
TopPharmComp %>% 
  # remove characters
  dplyr::mutate(year = str_remove_all(string = year, pattern = "\\D"),
                year = str_sub(string = year, start = 1L, end = 4L),
                # make numeric
                year = as.integer(year)) -> TopPharmComp
```

## `company_name` 

We want the name of the company (and only it's name) in it's own column. 

### `distinct()`

The first thing we will do is take a look at the `distinct()` of companies in this dataset. 

```{r distinct}
TopPharmComp %>% 
  distinct(company)
```

We can see a lot of text garbage (`\n` and `\\`), so we'll remove this first and create our `company_name` column 

### Remove newline characters

We can remove newline characters with `str_remove_all()` from `stringr`. 

```{r company_name}
TopPharmComp %>% 
  mutate(company_name = str_remove_all(company, "\n|//")) %>% 
  distinct(company, company_name)
```

### Remove numeric bracket contents

This looks good-now we want to remove anything after the first open bracket `[`. I can Google "matching string contents in brackets in R" and get this  [answer](https://stackoverflow.com/questions/50462907/regex-in-r-to-match-strings-in-square-brackets), which I can adapt to include `\\d` for all digits. 

```{r str_remove_all-brackets}
TopPharmComp %>% 
  mutate(
    company_name = str_remove_all(company, "\n|//"),
    company_name = str_remove_all(company_name, "(\\[\\d\\])")
    ) %>% 
  distinct(company, company_name)
```

### Replace non-numeric bracket contents

Now we want to replace all of the `[P]` or `[B]` with a space and a dash, which I can do with `str_replace_all()` and `\\D` for all non-digits.

```{r str_replace_all}
TopPharmComp %>% 
  mutate(
    company_name = str_remove_all(company, "\n|//"),
    company_name = str_remove_all(company_name, "(\\[\\d\\])"),
    company_name = str_replace_all(company_name, "(\\[\\D\\])", " - ")
    ) %>% 
  distinct(company, company_name)
```

### Separate `company_name` from `stock_info` 

Here we `separate` the `company_name` column into two new columns, 

```{r check-company_type}
TopPharmComp %>% 
  mutate(
    company_name = str_remove_all(company, "\n|//"),
    company_name = str_remove_all(company_name, "(\\[\\d\\])"),
    company_name = str_replace_all(company_name, "(\\[\\D\\])", " - ")
    ) %>%  
  # separate company into 2 columns
  separate(col = company_name, into = c("company_name", "stock_info"), 
           sep = " - ", 
           remove = FALSE) %>% 
  distinct(company, company_name, stock_info)
```

These look are correct, so we can assign `company_name` to `TopPharmComp`.

```{r assign-company_name}
TopPharmComp <- TopPharmComp %>% 
  mutate(
    company_name = str_remove_all(company, "\n|//"),
    company_name = str_remove_all(company_name, "(\\[\\d\\])"),
    company_name = str_replace_all(company_name, "(\\[\\D\\])", " - ")
    ) %>%  
  # separate company into 2 columns
  separate(col = company_name, into = c("company_name", "stock_info"), 
           sep = " - ", 
           remove = FALSE)
TopPharmComp
```

## `stock_exch` & `stock_id` 

Now we have the `stock_exch` and `stock_id` stored in `stock_info`--we will create two new columns for each piece of data. 

### `stock_exch`

We will use the `company_etc` column to create `stock_exch` and `stock_id`. 

```{r check-stock_exch-stock_id}
TopPharmComp %>% 
  # first separate the columns
    separate(col = stock_info, 
             into = c("stock_exch", "stock_id"), 
             sep = ":", remove = FALSE) %>% 
  # check our work
  select(company,
           contains("stock")) %>% 
  distinct()
```


### `stock_id`

This looks good, but we also need to make sure we trim the white space from `stock_id` (after the colon). The `sep` argument helps us a lot here, so we can create two new variables with minimal steps! We'll assign to `TopPharmComp`.

```{r assign-stock_exch-stock_id}
TopPharmComp <- TopPharmComp %>% 
  # first separate the columns
    separate(col = stock_info, 
             into = c("stock_exch", "stock_id"), 
             sep = ":", remove = FALSE) 
TopPharmComp
```

## `largest_market_cap_us_bil` & `largest_market_cap_date` 

Here we'll use `dplyr::rename()` to change `largest_market_cap_us_bil` to `_original` so we can check our work with `remove = FALSE`.

### separate market cap & date

We want to separate the `largest_market_cap_original` column into `largest_market_cap_us_bil` and `largest_market_cap_date` on the first open parenthesis. 

```{r check-largest_market_cap_date}
TopPharmComp %>% 
  # rename column to largest_market_cap_original
  dplyr::rename(largest_market_cap_original = largest_market_cap_us_bil) %>% 
  # separate into two columns
  separate(col = largest_market_cap_original, 
           into = c('largest_market_cap_us_bil', 'largest_market_cap_date'),
           sep = "\\(", 
           remove = FALSE) %>% 
  distinct(largest_market_cap_original, largest_market_cap_us_bil, 
    largest_market_cap_date)
```


We see some messages and warnings (which we will ignore), but this looks like it's creating the two columns we want. We just need to clean up the dates a bit. 

### Remove parenthesis 

Now we remove the remaining parenthesis. 

```{r str_remove_all-paren}
TopPharmComp %>% 
  # rename column to largest_market_cap_original
  dplyr::rename(largest_market_cap_original = largest_market_cap_us_bil) %>% 
  # separate into two columns
  separate(col = largest_market_cap_original, 
           into = c('largest_market_cap_us_bil', 'largest_market_cap_date'),
           sep = "\\(", 
           remove = FALSE) %>% 
  # remove parenthesis from date
  mutate(largest_market_cap_date = 
           str_remove_all(largest_market_cap_date, pattern = "\\)$")) %>% 
  # check our work
    distinct(largest_market_cap_original, largest_market_cap_us_bil, 
    largest_market_cap_date)
```

### Format dates and numbers 

Now we have the month and year separated into `largest_market_cap_date`, but we should convert this into a date and the `largest_market_cap_us_bil` column into a number. 

```{r format-date-and-number}
TopPharmComp %>% 
  # rename column to largest_market_cap_original
  dplyr::rename(largest_market_cap_original = largest_market_cap_us_bil) %>% 
  # separate into two columns
  separate(col = largest_market_cap_original, 
           into = c('largest_market_cap_us_bil', 'largest_market_cap_date'),
           sep = "\\(", 
           remove = FALSE) %>% 
  # remove parenthesis from date
  mutate(
    largest_market_cap_date = str_remove_all(largest_market_cap_date, pattern = "\\)$"),
    # date
    largest_market_cap_date = lubridate::my(largest_market_cap_date), 
    # number
    largest_market_cap_us_bil = as.numeric(largest_market_cap_us_bil)
    ) %>% 
    # check our work
    distinct(largest_market_cap_original, largest_market_cap_us_bil, 
    largest_market_cap_date)
```

These look good--we can assign to `TopPharmComp`

```{r assign-largest_market_cap_date-largest_market_cap_us_bil}
 TopPharmComp <- TopPharmComp %>% 
  # rename column to largest_market_cap_original
  dplyr::rename(largest_market_cap_original = largest_market_cap_us_bil) %>% 
  # separate into two columns
  separate(col = largest_market_cap_original, 
           into = c('largest_market_cap_us_bil', 'largest_market_cap_date'),
           sep = "\\(", 
           remove = FALSE) %>% 
  # remove parenthesis from date
  mutate(
    largest_market_cap_date = str_remove_all(largest_market_cap_date, pattern = "\\)$"),
    # date
    largest_market_cap_date = lubridate::my(largest_market_cap_date), 
    # number
    largest_market_cap_us_bil = as.numeric(largest_market_cap_us_bil)
    )
TopPharmComp
```

## `market_cap_us_bil` 

The `market_cap_us_bil` column has a few extra characters from the Wikipedia table (they use `-` as missing). We know how to remove this with str_remove_all() and we can format this as a number with `as.numeric()`

```{r market_cap_us_bil}
TopPharmComp %>% 
  mutate(
    market_cap_us_bil = str_remove_all(market_cap_us_bil, "-"), 
    market_cap_us_bil = as.numeric(market_cap_us_bil)
    ) %>% 
  distinct(market_cap_us_bil)
```

Looks good! Assign it to `TopPharmComp`

```{r assign-market_cap_us_bil}
TopPharmComp <- TopPharmComp %>% 
  mutate(
    market_cap_us_bil = str_remove_all(market_cap_us_bil, "-"), 
    market_cap_us_bil = as.numeric(market_cap_us_bil)
    )
TopPharmComp
```

## `ranking` & `company_type` 

The `rank` variable will come in handy for plotting, but we will rename it as `ranking`. We also want to differentiate between `[B]` – Biotechnology company and `[P]` – Pharmaceutical company.

### Remove extra numbers from rank

Here we can use `str_remove_all()` to remove the extra numbers in `rank` and create the `ranking` variable.

```{r check-ranking}
TopPharmComp %>% 
  dplyr::mutate(ranking = 
          stringr::str_remove_all(string = rank, pattern = " \\d$")) %>% 
  select(contains("rank"))
```

We assign this to `TopPharmComp`.

```{r assign-ranking}
TopPharmComp <- TopPharmComp %>% 
  dplyr::mutate(ranking = 
          stringr::str_remove_all(string = rank, pattern = " \\d$")) 
TopPharmComp
```

### Identify company type 

Now we combine `case_when()` with `str_detect()` for the company type. 

```{r company_type}
TopPharmComp %>% 
  mutate(
    company_type = case_when(
    str_detect(company, "[P]") ~ "Pharmaceutical company",
    str_detect(company, "[B]") ~ "Biotechnology company",
    TRUE ~ NA_character_
    )) %>% 
  distinct(company, company_type)
```


We assign `company_type` to `TopPharmComp`.

```{r assign-company_type}
TopPharmComp <- TopPharmComp %>% 
  mutate(
    company_type = case_when(
    str_detect(company, "[P]") ~ "Pharmaceutical company",
    str_detect(company, "[B]") ~ "Biotechnology company",
    TRUE ~ NA_character_
    ))
TopPharmComp
```


## Reorganize columns 

The data still contains all of the original columns, but now we're confident we can remove these. 

### Preview wrangled data 

Let's take a look with `glimpse()` to see how these are formatted. 

```{r TopPharmComp-glimpse}
# set width
options(width = 78)
# view transposed data
TopPharmComp %>% glimpse()
```

### Select new columns

We'll do this by only selecting the new columns. Notice we change the name of `year` to `market_cap_year`, 

```{r preview-TopPharmComp-cols}
select(TopPharmComp, 
  ranking,
  company_name, 
  company_type,
  largest_market_cap_us_bil, 
  largest_market_cap_date,
  market_cap_us_bil, 
  market_cap_year = year, 
  stock_exch, 
  stock_id)
```

We'll assign this to `TopPharmComp` 

```{r assign-TopPharmComp-cols}
TopPharmComp <- select(TopPharmComp, 
  ranking,
  company_name, 
  company_type,
  largest_market_cap_us_bil, 
  largest_market_cap_date,
  market_cap_us_bil, 
  market_cap_year = year, 
  stock_exch, 
  stock_id)
TopPharmComp
```


## (Optional) convert ranking to factor   

We want to keep the order of `ranking`, so another option is to convert this to a factor. 

### `ordered = TRUE`

If we format `ranking` as an ordered factor, we don't need to supply any `levels` because they are already numeric. 

```{r factor-ranking}
TopPharmComp <- TopPharmComp %>% 
  mutate(
    ranking = factor(ranking, ordered = TRUE))
# check
TopPharmComp %>% arrange(ranking)
```

# Export data 

We now have a dataset we can use for modeling or visualizations! We'll export this to the `data/wikipedia-data/` folder.

```{r export-TopPharmComp}
toppharm_comp_export_file <- paste0("data/wikipedia-data/", lubridate::today(), "-TopPharmComp.csv")
# toppharm_comp_export_file
readr::write_csv(x = TopPharmComp, file = toppharm_comp_export_file)
fs::dir_tree("data/wikipedia-data/")
```

