
<!--


It's hard to tell if the code above is working because we're still seeing all the columns in `big_starwars`. Reduce the output by adding `dplyr::select()` to the pipeline. 

```{r filter-sol-01a}
big_starwars |> 
  dplyr::filter(species == "Droid") |> 
  dplyr::select(species)
```


The code below `big_starwars` only returns data from the `homeworld`s of `Tatooine` and `Alderaan`:

```{r filter-sol-02}
big_starwars |> 
  dplyr::filter(homeworld %in% c("Tatooine", "Alderaan")) |>
  dplyr::select(homeworld)
```

I don't want to try and count each row to verify if it's **only** returning the two `homeworld`s of interest, so we can use `dplyr::count()` to verify the returned categories.

```{r filter-sol-02a}
big_starwars |>
  dplyr::filter(homeworld %in% c("Tatooine", "Alderaan")) |>
  dplyr::count(homeworld)
```

The code below only returns data from `big_starwars` where `species` is `Gungan`, `Hutt`, or `Wookie`:

```{r filter-sol-03}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  select(species)
```

We can use `dplyr::count()` with multiple variables to get the specific number of rows for combinations of values. 

For example, the code below gives us the count of the three categories of `species` above by their `species_class` and `homeworld`:

```{r filter-sol-03a}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  dplyr::count(species_class, homeworld)
```

We can include `sort = TRUE` to arrange the `n` column descending: 

```{r filter-sol-03b}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  dplyr::count(species_class, homeworld, sort = TRUE)
```

It's important to note that `dplyr::count()` returns the variable's distinct counts for each level or value, so it's less helpful if you're using it with a numeric variable (or a variable with unique values per row). 

For example, the code below returns data where `mass` is greater than `180` and `species` does not include `"Droid"`.   

```{r filter-sol-04}
big_starwars |> 
  dplyr::filter(mass > 100 & species != "Droid") |> 
  select(mass, species)
```


If we swap `select()` for `count()`, we get an additional `n` column (but all equal `1`), so it adds no new information. 

```{r filter-sol-04a}
big_starwars |> 
  dplyr::filter(mass > 100 & species != "Droid") |> 
  count(mass, species)
```


# Arranging data 

`dplyr::arrange()` sorts variables on their content, numeric or character. 

The code below sorts `big_starwars` according to the `hair_color` and views to top (`head(10)`) and bottom (`tail(10)`) ten rows of the result:

```{r arrange-sol-01a}
big_starwars |> 
  dplyr::arrange(hair_color) |> 
  select(hair_color) |> 
  head(10)
```

```{r arrange-sol-01b}
big_starwars |> 
  dplyr::arrange(hair_color) |> 
  select(hair_color) |> 
  tail(10)
```

Note that the missing values are sorted to the bottom of the returned column.

The code below filters the `big_starwars` data to only rows where the `species_class` is `amphibian` and `artificial`, then sorts the results by `species_class` and `mass`, descending:

```{r arrange-sol-02}
big_starwars |> 
  dplyr::filter(species_class %in% c("amphibian", "artificial")) |>
  dplyr::arrange(species_class, desc(mass)) |> 
  select(species_class, mass)
```

This is handy when you want to view the output sorted by multiple levels of a categorical variable. 

# Mutating columns 

`mutate()` can create new columns, or change existing columns.  

The code below creates a new `bmi` column in `big_starwars` using `mass` and `height`:

```{r mutate-sol-01}
big_starwars |>
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2))
```

Note the use of parentheses here. 

When we're creating new variables with `dplyr::mutate()`, the new variable gets pushed to the end of the dataset. We can view this column by dropping the `last_col()` function inside `select()`

```{r mutate-sol-01a}
big_starwars |> 
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2)) |>
  dplyr::select(last_col())
```



The code below creates a new `bmi25_ind` variable that splits the new `bmi` variable into above or below `25`, then filters the results to `Hutt` and `Wookie`, then counts and sorts the results: 

```{r mutate-sol-02}
big_starwars |>
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2),
    bmi25_ind = bmi > 25) |>
  dplyr::filter(species %in% c("Hutt", "Wookie")) |>
  dplyr::count(species_class, bmi25_ind, sort = TRUE)
```

# What we've covered 

1. Pipes:  (`|>`  and `%>%>`)     
2. Selecting columns with `select()` and [select helpers](https://tidyselect.r-lib.org/index.html)  
3. Filtering rows and `count()`ing the results   
4. Arranging (sorting) data based on their values   
5. Creating new columns with `mutate()`   