---
title: "FA22 - JOUR 301"
description: |
  Welcome to Data Journalism 301!
site: distill::distill_website
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Learn more about creating websites with Distill at:
# https://rstudio.github.io/distill/website.html

# Learn more about publishing to GitHub Pages at:
# https://rstudio.github.io/distill/publish_website.html#github-pages

```

This is the home page for JOUR 301, Data Journalism. All course materials will be posted on this site.

Instructor: [Martin Frigaard](https://www.martinfrigaard.io/)  
Meeting times: Mon 5:00pm to 7:50pm  
Syllabus: [click here](syllabus.html)  
Upcoming lectures and assignments: [click here](schedule.html)  

## Software requirements

For students enrolled in this course, you will need a computer with a working web browser. Computational requirements for using R & RStudio are described [here](https://support.rstudio.com/hc/en-us/articles/201853926-RStudio-IDE-System-Recommendations), but most modern laptops will run both programs without issue. 

### R

Download a recent version of R [here](https://cloud.r-project.org/)

### RStudio

Download a recent version of RStudio [here](https://rstudio.com/products/rstudio/download/#download)

### Installation Instructions

See instructions for installing R and RStudio are available here. 

### R packages for this course 

This course required the use of two packages on Github: 

1. [`goodenuffR`](https://mjfrigaard.github.io/goodenuffR/)  

2. [`sheetcheatR`](https://mjfrigaard.github.io/sheetcheatR/)

Install these packages with the following lines of code in your R console: 

```{r eval = FALSE, echo = TRUE}
install.packages("remotes")
remotes::install_github("mjfrigaard/goodenuffR")
remotes::install_github("mjfrigaard/sheetcheatR")
```


### Additional R packages used in this course

`dplyr`, `tidyr`, `fs`, `forcats`, `stringr`, `rmarkdown`, `lubridate`, 
`ggplot2`, `cowplot`, `gapminder`, `GGally`, `gganimate`, `ggiraph`, `ggdendro`,
`ggdist`, `ggforce`, `ggplot2movies`, `ggrepel`, `ggridges`, `ggthemes`, 
`gifski`, `glue`, `knitr`, `learnr`, `naniar`, `margins`, `MASS`, `Matrix`, 
`nycflights13`, `palmerpenguins`, `patchwork`, `rgdal`, `rmarkdown`, `umap`, 
`xaringan`, `distill`

Install these packages with the following lines of code in your R console:

```{r eval = FALSE, echo = TRUE}
install.packages(c("dplyr", "tidyr", "fs", "forcats", "stringr", "rmarkdown",
  "lubridate", "ggplot2", "cowplot", "gapminder", "GGally", "gganimate", 
  "ggiraph", "ggdendro", "ggdist", "ggforce", "ggplot2movies", "ggrepel", 
  "ggridges", "ggthemes", "gifski", "glue", "knitr", "learnr", "naniar", 
  "margins", "MASS", "Matrix", "nycflights13", "palmerpenguins", "patchwork",
  "rgdal", "rmarkdown", "umap", "xaringan", "distill"
))
```

