---
title: "Google trends (example EDA with goodenuffR)"
author: "Your Name"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---

```{r , include=FALSE}
library(knitr)
library(tidyverse)
library(janitor)
# base options ----
base::options(
  tibble.print_max = 25,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
# knitr chunk options ----
knitr::opts_chunk$set(
  echo = TRUE, # show/hide all code
  # results = "hide", # hide/show results
  tidy = FALSE, # cleaner code printing
  comment = "#> ", # better console printing
  eval = TRUE, # turn this to FALSE stop code chunks from running
  message = TRUE, # show messages
  fig.width = 7, # figure width
  fig.height = 5, # figure height
  warning = FALSE, # show warnings
  size = "small", # size of the text
  fig.path = "img/" # location of figures
) 
# knitr knit settings ----
knitr::opts_knit$set(
  width = 78
)
```

## Motivation

This document will cover an example EDA project using the `gtrendsR` package to gather Google trend data.

Below we install and load the `gtrendsR` package...

```{r gtrendsR}
# devtools::install_github("PMassicotte/gtrendsR")
library(gtrendsR)
```

## Code/data files & folders {.tabset}

We will use the handy `goodenuffR::goodenuff_code()` and `goodenuffR::goodenuff_data()` functions to create the code files. 

### The code folder

Create the folder/files:

```{r goodenuff_collab}
goodenuffR::goodenuff_code()
```

Verify it's there:

```{r code-folder}
fs::dir_tree("code")
```

### The data folder


Create the folder/files:

```{r goodenuff_data}
goodenuffR::goodenuff_data() 
```

Verify it's there:

```{r code-folder}
fs::dir_tree("data")
```

## Import data {.tabset}

We will collect Google trends for the term 'metaverse'. Facebook recently announced this name change for their company. What is the 'metaverse'? No one [really knows](https://arstechnica.com/gaming/2021/11/everyone-pitching-the-metaverse-has-a-different-idea-of-what-it-is/). 

### `gtrends()` data

The code below will import the trend data over the past 12 months and store it in `MetaverseTrendsRaw`.

```{r MetaverseTrendsRaw}
MetaverseTrendsRaw <- gtrendsR::gtrends(keyword = "metaverse", time = "today 12-m", gprop = "web")
names(MetaverseTrendsRaw)
```

### Import script

We want to add all the previous code to our `01-import.R` script.

```{r 01-import.R, eval=FALSE}
#=====================================================================#
# This is code to create: import google trends
# Authored by and feedback to: mjfrigaard
# MIT License
# Version: 1.0
#=====================================================================#


# packages  ---------------------------------------------------------------
library(tidyverse)
library(janitor)
# devtools::install_github("PMassicotte/gtrendsR")
library(gtrendsR)


# import google trends ----------------------------------------------------
MetaverseTrendsRaw <- gtrendsR::gtrends(keyword = "metaverse",
  time = "today 12-m", gprop = "web")
```

### Export `Raw` data 

In keeping with good practices, we're going to export the raw data to the `data/raw/` folder. This is a little more complicated because `gtrendsR` returned a list, but we can separated these out into individual datasets.

```{r new-names}
paste0("meta_", names(MetaverseTrendsRaw), "_raw")
```


```{r separate-data}
meta_interest_over_time_raw <- MetaverseTrendsRaw$interest_over_time
meta_interest_by_country_raw <- MetaverseTrendsRaw$interest_by_country
meta_interest_by_region_raw <- MetaverseTrendsRaw$interest_by_region
meta_interest_by_dma_raw <- MetaverseTrendsRaw$interest_by_dma
meta_interest_by_city_raw <- MetaverseTrendsRaw$interest_by_city
meta_related_topics_raw <- MetaverseTrendsRaw$related_topics
meta_related_queries_raw <- MetaverseTrendsRaw$related_queries
```

```{r export-raw}
readr::write_csv(
  x = meta_interest_over_time_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_over_time_raw.csv"))

readr::write_csv(
  x = meta_interest_by_country_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_country_raw.csv"))

readr::write_csv(
  x = meta_interest_by_dma_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_dma_raw.csv"))

readr::write_csv(
  x = meta_interest_by_city_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_city_raw.csv"))

readr::write_csv(
  x = meta_related_topics_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_related_topics_raw.csv"))

readr::write_csv(
  x = meta_related_queries_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_related_queries_raw.csv"))

```

## Visualize 

We can supply this object directly to the `plot()` function and see the trends for metaverse across the world over time. 

```{r}
plot(MetaverseTrendsRaw)
```

### interest_over_time

### interest_by_country

### interest_by_region



## Data folder 

- The data folder also has a `raw` sub-folder.

## The doc folder

```{r goodenuff_doc}
goodenuffR::goodenuff_doc() 
```

> The `doc` directory contains two text files written in Markdown, one containing a running lab notebook describing various ideas for the project and how these were implemented and the other containing a running draft of a manuscript describing the project findings.

### The CHANGELOG.txt

The `CHANGELOG.md` file is for documenting "*dated notes about changes to the project in this file in reverse chronological order (i.e., most recent first). This file is the equivalent of a lab notebook, and should contain entries like those shown below.* This can also referred to as the `NEWS.md` file. 

```
...make dated notes about changes to the project in this file in reverse chronological order (i.e., most recent first). This file is the equivalent of a lab notebook, and should contain entries like those shown below.

## 2016-04-08

* Switched to cubic interpolation as default.

* Moved question about family's TB history to end of questionnaire.

## 2016-04-06

* Added option for cubic interpolation.

* Removed question about staph exposure (can be inferred from blood test results).
```

### The manuscript.md

Draft of your manuscript.

```
# Manuscript

...a running draft of a manuscript describing the project findings
```

### The notebook.md

For your brilliant ideas

```
# Notebook

...a running lab notebook describing various ideas for the project and how these were implemented


```


## The results folder

```{r goodenuff_results}
goodenuffR::goodenuff_results() 
```

These are visualizations, model outputs, etc. It has three sub-folders:

```
results
    ├── figures
    ├── manuscript
    └── tables
```

## Project folder structure

Below is a nice overview of the accompanying files. 

```r
├── CITATION
├── LICENSE
├── code
│   ├── 01-import.R
│   ├── 02-tidy.R
│   ├── 03-wrangle.R
│   ├── 04-visualize.R
│   ├── 05-model.R
│   ├── 06-communicate.R
│   └── runall.R
├── data
│   ├── README.md
│   └── raw
├── docs
│   ├── changelog.txt
│   ├── manuscript.md
│   └── notebook.md
├── my-new-project.Rproj
├── requirements.txt
└── results
    ├── figures
    ├── manuscript
    └── tables
```

