---
title: "Google trends (example EDA)"
date: "`r Sys.Date()`"
description: |
  "An example using gtrendsR package."
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2

always_allow_html: true

---

```{r setup, include=FALSE}
library(knitr)
library(tidyverse)
library(janitor)
# base options ----
base::options(
  tibble.print_max = 25,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
# knitr chunk options ----
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  dpi = 320,
  out.width = "80%",
  fig.width = 6,
  fig.asp = 0.618,
  fig.retina = 2,
  fig.align = "center",
  fig.show = "hold"
  )
# knitr knit settings ----
knitr::opts_knit$set(
  width = 78
)
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/google-trends/')
```

## Motivation

This document will cover an example EDA project using the `gtrendsR` package to gather Google trend data.

Below we install and load the `gtrendsR` package...

```{r gtrendsR}
# devtools::install_github("PMassicotte/gtrendsR")
library(gtrendsR)
```

## Import data {.tabset}

We will collect Google trends for the term 'metaverse'. Facebook recently announced this name change for their company. What is the 'metaverse'? No one [really knows](https://arstechnica.com/gaming/2021/11/everyone-pitching-the-metaverse-has-a-different-idea-of-what-it-is/). 

### `gtrends()` data

The code below will import the trend data over the past 12 months and store it in `MetaverseTrendsRaw`.

```{r load-MetaverseTrendsRaw, eval=TRUE, echo=FALSE}
MetaverseTrendsRaw <- readr::read_rds(file = "data/2022-10-18-MetaverseTrendsRaw.rds")
```


```{r MetaverseTrendsRaw, eval=FALSE}
MetaverseTrendsRaw <- gtrendsR::gtrends(keyword = "metaverse", 
  time = "today 12-m", gprop = "web")
```

```{r names-MetaverseTrendsRaw}
names(MetaverseTrendsRaw)
```

## Export `Raw` data 

In keeping with good practices, we're going to export the raw data to the `data/raw/` folder. This is a little more complicated because `gtrendsR` returned a list, but we can separated these out into individual datasets.

### Create raw data names 

```{r new-names}
paste0("meta_", names(MetaverseTrendsRaw), "_raw")
```

### Separate from list 

```{r separate-data}
meta_interest_over_time_raw <- MetaverseTrendsRaw$interest_over_time
meta_interest_by_country_raw <- MetaverseTrendsRaw$interest_by_country
meta_interest_by_region_raw <- MetaverseTrendsRaw$interest_by_region
meta_interest_by_dma_raw <- MetaverseTrendsRaw$interest_by_dma
meta_interest_by_city_raw <- MetaverseTrendsRaw$interest_by_city
meta_related_topics_raw <- MetaverseTrendsRaw$related_topics
meta_related_queries_raw <- MetaverseTrendsRaw$related_queries
```

### Export to `data/raw/`

```{r export-raw}
readr::write_csv(
  x = meta_interest_over_time_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_over_time_raw.csv"))

readr::write_csv(
  x = meta_interest_by_country_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_country_raw.csv"))

readr::write_csv(
  x = meta_interest_by_dma_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_dma_raw.csv"))

readr::write_csv(
  x = meta_interest_by_city_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_interest_by_city_raw.csv"))

readr::write_csv(
  x = meta_related_topics_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_related_topics_raw.csv"))

readr::write_csv(
  x = meta_related_queries_raw, 
  file = 
    paste0("data/raw/", base::noquote(lubridate::today()), 
      "-meta_related_queries_raw.csv"))

fs::dir_tree("data/raw/")
```

## Visualize 

We can supply this object directly to the `plot()` function and see the trends for metaverse across the world over time. 

```{r}
plot(MetaverseTrendsRaw)
```

