---
title: "My Storyboard"
output: 
  flexdashboard::flex_dashboard:
    theme: default
    storyboard: true
    self_contained: true

---

```{r setup, include=FALSE}
library(tidyverse)
# create www folder
fs::dir_create("www/")
# options for chunks
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "www/",
  dpi = 320,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5.5,
  fig.retina = 2
  )
# tibble printing
options(
  pillar.print_max = 10, 
  pillar.print_min = 10)
```


### Background {data-commentary-width=350}

The Palmer Penguins datasets comes from the [`palmerpenguins` package](https://allisonhorst.github.io/palmerpenguins/). You can install it using the code below:

```{r install_palmerpenguins, eval=FALSE}
install.packages("palmerpenguins")
# or 
# install.packages("remotes")
remotes::install_github("allisonhorst/palmerpenguins")
```

*** 

Original paper on Palmer Penguins: 

*Ecological Sexual Dimorphism and Environmental Variability within a Community of Antarctic Penguins (Genus Pygoscelis)*  
Kristen B. Gorman, Tony D. Williams, William R. Fraser. [PloS ONE](https://doi.org/10.1371/journal.pone.0090081)

### Import {data-commentary-width=350}

Code for obtaining data...

```{r}
```

*** 

Information about the dataset...

### Tidy/Wrangle {data-commentary-width=300}

Code for tidying/wrangling data

```{r}
```

*** 

More details on data...

### Visualize {data-commentary-width=400}

Code for visualizing the data

```{r}

```

***

More details on visualizations 

### Future Work {data-commentary-width=300}

