---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Week 2: Intro to the goodenuffR package"
author: "<Your Name>"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")
```

We'll be using R Markdown for most of the assignments in this course. As you'll discover, R Markdown is a powerful tool because it can combine prose/text, R code chunks and outputs (like figures and tables).

In this assignment, we're going to cover how to use the `goodenuffR` package to organize your folders and files in this course. 

# Installtion

Run the code chunk below to install and load `goodenuffR`. 

```{r install}
install.packages("remotes")
remotes::install_github("mjfrigaard/goodenuffR")
```

Follow the instructions on using `goodenuffR` on the [package website.](https://mjfrigaard.github.io/goodenuffR/).



