---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Week 2: Intro to the goodenuffR package"
author: "<Your Name>"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")
```

We'll be using R Markdown for most of the assignments in this course. As you'll discover, R Markdown is a powerful tool because it can combine prose/text, R code chunks and outputs (like figures and tables).

In this assignment, we're going to cover how to use the `goodenuffR` package to organize your folders and files in this course. 

To get credit for this assignment, you'll need to 

1) successfully install the `goodenuffR` package,  
2) create a course folder  
3) create a homework assignment folder  
4) use the `goodenuffR` to create a project folder for this assignment following the example on the course website: https://mjfrigaard.github.io/FA22-JOUR301/exercises/goodenuffR.html
5) run the code chunk in the **Folder Tree** section to confirm the project was built correctly. 

Save this R Markdown file and name it with the following conventions: 

`YYYY-MM-DD-LASTNAME-wk3-rmarkdown-in-practice.Rmd`

*Replace the date (`YYYY-MM-DD`) and enter your last name (`LASTNAME`).*

Place this file in the top-level folder of the project: 

**Edit only <Your Name> above and below this line.**

---------------------------------------------

# Installtion

Copy the code below and run it in the console to install and load the `goodenuffR` package.  

```{r install, eval=FALSE}
install.packages("remotes")
remotes::install_github("mjfrigaard/goodenuffR")
```

# Folder Tree 

The code chunk below runs the `fs::dir_tree(".")` function, which prints the files and folders in given directory (and their structure). 

```{r dir_tree}
install.packages("fs", repos = "https://cloud.r-project.org")
library(fs)
fs::dir_tree(".")
```







