---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Week 5: ggplot2 part 1"
author: "<Your Name>"
output: html_document
---

```{r , include=FALSE}
library(tidyverse)
library(ipcwswitch)
library(skimr)
library(listviewer)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")
```

In this assignment, we're going to follow the steps in the first set of exercises for ggplot2. You can review the slides for this section [here](https://mjfrigaard.github.io/FA22-JOUR301/slides/ggp2-part-01.html#1), 

In all of the exercises below, the code chunks have been set to `eval=FALSE`. In order to get credit for this assignment, you'll have to set each of them to `eval=TRUE` (or delete the `eval` argument).

When you've reached the end of this document, click on the "Knit" icon in R Studio and review the output.

**Edit only <Your Name> above and below this line.**

---------------------------------------------

# Project setup 

Follow the instructions for setting up your project with `goodenuffR` on the [course website.](https://mjfrigaard.github.io/FA22-JOUR301/exercises/goodenuffR.html) 

# Load the packages

The main packages we're going to use are `dplyr`, `tidyr`, and `ggplot2`. These are all part of the `tidyverse`, so we'll import this package below:

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

# Import

Import the data below

```{r import-AppleMobRaw, eval=FALSE}
AppleMobRaw <- readr::read_csv("https://bit.ly/3ALws1X")
```



