---
title: "Introduction to Data Manipulation"
description: |
  "*introduction to wrangling data*"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2

always_allow_html: true

params: 
  data_file: !r file.path("https://bit.ly/big-sw")
---

```{r setup, echo = FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>", 
                      error = TRUE,
                      eval = TRUE)
options(scipen = 100000000)
library(tidyverse)
library(skimr)
big_starwars <- read_csv("https://bit.ly/big-sw")
```


# Objectives 

This lesson outlines some common data manipulation tasks with `dplyr`: 

1. `select()` for columns  
2. `filter()` for rows   
3. `arrange()` for sorting   
4. `mutate()` for creating new columns  

# Materials 

View the slides on the [course website.](https://mjfrigaard.github.io/FA22-JOUR301/slides/dplyr-part-01.html) 

Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

# Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

# Namespacing 

Namespacing is what we're doing when we using the `pkg::fun()` syntax. We do this because sometimes two different packages will use the same name for a function (after all, there are only so many names!).

This is why we see the following message when we load the `tidyverse` package:

```r
── Conflicts ───────────────────────────────────── tidyverse_conflicts
✖ dplyr::filter() masks stats::filter()
```

If we want to use the `filter()` function from the `stats` package, we'd need to use `stats::filter()`. 

If we use the `package::function()` syntax, we don't need to load the package with `library(package)` (as long as we've previously installed it with (`install.packages("package")`)).

# Import data

We'll cover two methods for importing data into RStudio. 

1. importing data from a package    
2. Importing data using `params` in the YAML header  

We have the path to the `big_starwars` data stored in our `params`, but we will also go over how to load a smaller version of this dataset from `dplyr::starwars`. 

## Importing data from a package

Below we import the `starwars` dataset from the `dplyr` package. This is similar to the way we'd access a function from a package (`dplyr::starwars`).

```{r import-pkg_starwars}
pkg_starwars <- dplyr::starwars
glimpse(pkg_starwars)
```

## Importing data using `params`

We have the `params` list from our YAML header, which we can also use to import the data. 

```{.yaml}
params: 
  data_file: !r file.path("https://bit.ly/big-sw")
```

```{r import-params}
big_starwars <- readr::read_csv(params$data_file)
glimpse(big_starwars)
```

Check to see which columns are different using `names()` and `setdiff()`: 

```{r identical-starwars}
setdiff(x = names(big_starwars), 
        y = names(pkg_starwars))
```

These are the columns in `big_starwars` that aren't in `pkg_starwars`. 

# `dplyr` verbs 

These exercises cover four common `dplyr` verbs for data manipulation: 

- `select`: works on **columns**/**variables**
- `filter`: works on **rows**/**observations**  
- `arrange`: sorts values (either ascending or descending (with `desc()`))
- `mutate`: creates new columns/variables (or changes format/type/class of existing variables)

We'll also be introducing some additional `tidyselect` functions and filtering tips and tricks. 

# Pipes

Clearly written code makes it easier for machines *and* humans to read. The pipes (`|>` or `%>%` from [the `magrittr` package](https://magrittr.tidyverse.org/))  allows us to chain together multiple operations into functional 'pipelines'. 

The exercises below use the native pipe (`|>`), but either pipe is acceptable (`%>%`). Recall the pipes convert our code from: 

```{r eval=FALSE}
OUTPUT_01 <- function_01(OBJECT_01)
OUTPUT_02 <- function_02(OUTPUT_01)
```

To something easier to read:

```{r eval=FALSE}
OBJECT_01 |>  
  function_01() |> 
  function_02() -> OUTPUT_02
```

The pipe (`%>%`) takes an object that comes before it, and it drops the object into the function that comes after it.

In the example below, it actually wouldn't matter what order the functions are in--both would return the same result. 

```{r pipe-sol-01}
big_starwars |> 
  dplyr::filter(mass < 100) |>  
  dplyr::select(name, hair_color, species, homeworld)
```

Consider the following operations *without* using the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`  

```{r pipe-sol-02}
# 1) 
x <- c(3, 7, 12)
# 2) 
mean_x <- mean(x)
# 3 
sqrt(mean_x)
```

This returns a vector, not a `tibble`.

Now let's re-write this code *with* the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`

```{r pipe-sol-03}
c(3, 7, 12) |> 
  mean() |> 
  sqrt()
```

Note that we can create a pipeline without even creating an object.

# Selecting columns 

We'll start with some exercises that demonstrate how to pick a subset of columns from your dataset using the `select()` function (and helpers!).

The general form of `select()` is below: 

```{r select-args, eval=FALSE}
dplyr::select(.data = , ... = )
```

The code below selects the `name` and `homeoworld` columns:

```{r select-sol-01}
big_starwars |> 
  dplyr::select(name, homeworld)
```

The code below selects only the columns starting with the letter `h`. 

```{r select-sol-02}
big_starwars |> 
  dplyr::select(starts_with("h"))
```

The code select *only* numeric columns starting with the letter `h`

```{r select-sol-03}
big_starwars |> 
  dplyr::select(starts_with("h")) |> 
  dplyr::select(where(is.numeric))
```


# Filtering rows 

Below are exercises for `dplyr::filter()`ing data. We're going to combine these with some of the `dplyr::select()` methods we covered above and in the slides. Recall that all filtering must be logical (i.e. evaluate the either `TRUE` or `FALSE`).

The code below only returns the droids in `big_starwars`:

```{r filter-sol-01}
big_starwars |> 
  dplyr::filter(species == "Droid") 
```


It's hard to tell if the code above is working because we're still seeing all the columns in `big_starwars`. Reduce the output by adding `dplyr::select()` to the pipeline. 

```{r filter-sol-01a}
big_starwars |> 
  dplyr::filter(species == "Droid") |> 
  dplyr::select(species)
```


The code below only returns data from the `homeworld`s of `Tatooine` and `Alderaan`:

```{r filter-sol-02}
big_starwars |> 
  dplyr::filter(homeworld %in% c("Tatooine", "Alderaan")) |>
  dplyr::select(homeworld)
```

I don't want to try and count each row to verify if it's **only** returning the two `homeworld`s of interest, so we can use `dplyr::count()` to verify the returned categories.

```{r filter-sol-02a}
big_starwars |>
  dplyr::filter(homeworld %in% c("Tatooine", "Alderaan")) |>
  dplyr::count(homeworld)
```

The code below only returns data from `big_starwars` where `species` is `Gungan`, `Hutt`, or `Wookie`:

```{r filter-sol-03}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  select(species)
```

We can use `dplyr::count()` with multiple variables to get the specific number of rows for combinations of values. 

For example, the code below gives us the count of the three categories of `species` above by their `species_class` and `homeworld`:

```{r filter-sol-03a}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  dplyr::count(species_class, homeworld)
```

We can include `sort = TRUE` to arrange the `n` column descending: 

```{r filter-sol-03b}
big_starwars |>
  dplyr::filter(species %in% c("Gungan", "Hutt", "Wookie")) |>
  dplyr::count(species_class, homeworld, sort = TRUE)
```

It's important to note that `dplyr::count()` returns the variable's distinct counts for each level or value, so it's less helpful if you're using it with a numeric variable (or a variable with unique values per row). 

For example, the code below returns data where `mass` is greater than `180` and `species` does not include `"Droid"`.   

```{r filter-sol-04}
big_starwars |> 
  dplyr::filter(mass > 100 & species != "Droid") |> 
  select(mass, species)
```


If we swap `select()` for `count()`, we get an additional `n` column (but all equal `1`), so it adds no new information. 

```{r filter-sol-04a}
big_starwars |> 
  dplyr::filter(mass > 100 & species != "Droid") |> 
  count(mass, species)
```


# Arranging data 

`dplyr::arrange()` sorts variables on their content, numeric or character. 

The code below sorts `big_starwars` according to the `hair_color` and views to top (`head(10)`) and bottom (`tail(10)`) ten rows of the result:

```{r arrange-sol-01a}
big_starwars |> 
  dplyr::arrange(hair_color) |> 
  select(hair_color) |> 
  head(10)
```

```{r arrange-sol-01b}
big_starwars |> 
  dplyr::arrange(hair_color) |> 
  select(hair_color) |> 
  tail(10)
```

Note that the missing values are sorted to the bottom of the returned column.

The code below filters the `big_starwars` data to only rows where the `species_class` is `amphibian` and `artificial`, then sorts the results by `species_class` and `mass`, descending:

```{r arrange-sol-02}
big_starwars |> 
  dplyr::filter(species_class %in% c("amphibian", "artificial")) |>
  dplyr::arrange(species_class, desc(mass)) |> 
  select(species_class, mass)
```

This is handy when you want to view the output sorted by multiple levels of a categorical variable. 

# Mutating columns 

`mutate()` can create new columns, or change existing columns.  

The code below creates a new `bmi` column in `big_starwars` using `mass` and `height`:

```{r mutate-sol-01}
big_starwars |>
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2))
```

Note the use of parentheses here. 

When we're creating new variables with `dplyr::mutate()`, the new variable gets pushed to the end of the dataset. We can view this column by dropping the `last_col()` function inside `select()`

```{r mutate-sol-01a}
big_starwars |> 
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2)) |>
  dplyr::select(last_col())
```



The code below creates a new `bmi25_ind` variable that splits the new `bmi` variable into above or below `25`, then filters the results to `Hutt` and `Wookie`, then counts and sorts the results: 

```{r mutate-sol-02}
big_starwars |>
  dplyr::mutate(bmi = mass / ((height / 100) ^ 2),
    bmi25_ind = bmi > 25) |>
  dplyr::filter(species %in% c("Hutt", "Wookie")) |>
  dplyr::count(species_class, bmi25_ind, sort = TRUE)
```

# What we've covered 

1. Pipes:  (`|>`  and `%>%>`)     
2. Selecting columns with `select()` and [select helpers](https://tidyselect.r-lib.org/index.html)  
3. Filtering rows and `count()`ing the results   
4. Arranging (sorting) data based on their values   
5. Creating new columns with `mutate()`   