---
title: "Introduction to Data Manipulation"
description: |
  "*introduction to wrangling data*"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2

always_allow_html: true

params: 
  data_file: !r dplyr::starwars
---

```{r , echo = FALSE, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, 
                      comment = "#>", 
                      error = TRUE)
library(tidyverse)
library(skimr)
original_starwars <- read_csv("https://bit.ly/3qgjqSC")
```


# Objectives 

This lesson outlines some common data manipulation tasks with `dplyr`: 

1. `select()` for columns  
2. `filter()` for rows   
3. `arrange()` for sorting   
4. `mutate()` for creating new columns  

# Materials 

View the slides on the [course website.](https://mjfrigaard.github.io/FA22-JOUR301/slides/dplyr-part-01.html) 

Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

# Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
```

# Import data

We'll cover two methods for importing data into RStudio. 

1. importing data from a package    
2. Importing data using `params` in the YAML header  

We have the path to the `starwars` data stored in our `params`, but we will also go over how to build this dataset from `dplyr::starwars`. 

## Importing data from the `dplyr` package

Below we import the `starwars` dataset from the `dplyr` package. This is similar to the way we'd access a function from a package (`dplyr::starwars`).

```{r import-pkg_starwars}
pkg_starwars <- dplyr::starwars
glimpse(pkg_starwars)
```

## Importing data using `params`

We have the `params` list from our YAML header, which we can also use to import the data. 

```{.yaml}
params: 
  data_file: !r dplyr::starwars
```

```{r import-params}
params_starwars <- params$data_file
glimpse(params_starwars)
```

Check to see if these two are identical: 

```{r identical-starwars}
identical(x = pkg_starwars, y = params_starwars)
```


# `dplyr` verbs 

This section covers four common `dplyr` verbs for data manipulation: 

- `select`: works on **columns**/**variables**
- `filter`: works on **rows**/**observations**  
- `arrange`: sorts values (either ascending or descending (with `desc()`))
- `mutate`: creates new columns/variables (or changes format/type/class of existing variables)

## Selecting columns 

We'll start with some exercises that demonstrate how to pick a subset of columns from your dataset using the `select()` function (and helpers!).

The general form of `select()` is below: 

```{r select-args, eval=FALSE}
select(.data = , ... = )
```


## Select exercise 1

Alter the code below to select just the `name` and `homeoworld` column:

```{r select-ex-01, eval=FALSE}
select(original_starwars, name, species, homeworld)
```


### solution

```{r select-sol-01}
select(original_starwars, name, homeworld)
```


### exercise 

Select only the columns starting with the letter `h`. 

```{r select-ex-02, eval=FALSE}
select(original_starwars, starts_with("_"))
```


### solution 

```{r select-sol-02}
select(original_starwars, starts_with("h"))
```

## Filtering rows {.tabset}

These are some additional exercises for `filter()`ing data. 

### exercise 

Change the code below so `original_starwars` only includes the droids. 

```{r filter-ex-01, eval=FALSE}
filter(original_starwars, species == "____")
```


### solution 

```{r filter-sol-01}
filter(original_starwars, species == "Droid")
```

### exercise 

Change the code below so `original_starwars` only includes data from the `homeworld`s of `Tatooine` and `Alderaan`

```{r filter-ex-02, eval=FALSE}
filter(original_starwars, 
       homeworld %in% c("________", "________"))
```

### solution 

```{r filter-sol-02}
filter(original_starwars, 
       homeworld %in% c("Tatooine", "Alderaan"))
```


## Arranging data {.tabset}

`arrange()` sorts variables on their content, numeric or character. 

### exercise  

Sort `original_starwars` according to the `hair_color`. 

```{r arrange-ex-01, eval=FALSE}
arrange(original_starwars, "____ _____")
```


### solution  

Note that the missing values are sorted to the bottom. 

```{r arrange-sol-01}
arrange(original_starwars, hair_color)
```


### exercise 

Sort `original_starwars` by `height` and `mass`, descending. 
```{r arrange-ex-02, eval=FALSE}
arrange(original_starwars, desc(______, ____))
```

### solution 

Including two variables is helpful if some of the values 'tie'. 

```{r arrange-sol-02}
arrange(original_starwars, desc(height, mass))
```

## Mutate columns {.tabset}

`mutate()` can create new columns, or change existing columns.  

### exercise 

Alter the code below to create a `bmi` column for `starwars` characters in `original_starwars`.

```{r mutate-ex-01, eval=FALSE}
mutate(original_starwars, 
       bmi = ____ / ((______ / 100) ^ 2))
```

### solution 

Note the use of parentheses here. 

```{r mutate-sol-01}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2))
```

### exercise 

Round the new `bmi` variable to 1 digit. 

```{r mutate-ex-02, eval=FALSE}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2),
       bmi = _____(___, digits = _))
```

### solution 

```{r mutate-sol-02}
mutate(original_starwars, 
       bmi = mass / ((height / 100) ^ 2),
       bmi = round(bmi, digits = 1))
```

# The pipe 

Clearly written code makes it easier for machines *and* humans to read. The pipe (`%>%`) from `magrittr` package allows us to chain together multiple operations into functional 'pipelines'. 

## Function pipelines {.tabset}

The pipe (`%>%`) takes an object that comes before it, and it drops the object into the function that comes after it.

### exercise 

Rewrite the code below to use the pipe 

```{r pipe-ex-01, eval=FALSE}
select(filter(
          original_starwars, mass < ___), 
              ____, hair_color, _______, homeworld)
```

### solution

This actually wouldn't matter what order it was in--both would return the same result. 

```{r pipe-sol-01}
original_starwars %>% 
  filter(mass < 100) %>% 
  select(name, hair_color, species, homeworld)
```


### exercise 

Perform the following operations *without* using the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`  

```{r pipe-ex-02, eval=FALSE}
# 1) 
_ <- c(_, _, __)
# 2) 
mean_x <- ____(x)
# 3 
sqrt(______)
```

### soluton 

This returns a vector, not a `tibble`.

```{r pipe-sol-02}
# 1) 
x <- c(3, 7, 12)
# 2) 
mean_x <- mean(x)
# 3 
sqrt(mean_x)
```

### exercise 

Perform the following operations *with* the pipe. 

1. create object `x` with three values, (`3`, `7`, `12`)    
2. get the `mean()` of `x`, and store it in `mean_x`   
3. take the squared root of `mean_x`

```{r pipe-ex-03, eval=FALSE}
c(_, _, __) %>% 
  ____() %>% 
  ____()
```


### solution 

Note that we can create a pipeline without even creating an object. 

```{r pipe-sol-03}
c(3, 7, 12) %>% 
  mean() %>% 
  sqrt()
```
