---
title: "Joining Data in R"
description: |
  "*Using `dplyr` to join data tables*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(inspectdf)
library(starwarsdb)
library(dm)
# create data folder
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5.5,
  fig.retina = 2
  )
# tibble printing
options(
  pillar.print_max = 10, 
  pillar.print_min = 10)
```

# Objectives 

This document outlines some common data joining tasks with the `dplyr` package. **If you've ever Googled 'how to perform [x] in R', you've probably discovered there are multiple ways to accomplish [x].** 

The methods I present here are not exhaustive, and I've tried to balance the trade-off between 1) solutions/functions that solve problems with the fewest number of keystrokes, and 2) code that's readable, explicit, and easy to follow.


## Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. We'll also be installing the `starwarsdb`, and `dm` packages. 

```{r packages, eval=FALSE}
# install.packages(c("tidyverse", "starwarsdb", "dm"))
library(tidyverse)
library(starwarsdb)
library(dm)
```

## Toy data 

We'll create the same example data from the slides 

```{r create-x}
# create X table
X <- tibble::tribble(
    ~A, ~B, ~C, 
    "a", "t", 1L,
    "b", "u", 2L,
    "c", "v", 3L)
X
```

```{r create-y}
# create Y table
Y <- tibble::tribble(
    ~A, ~B, ~D, 
    "a", "t", 3L,
    "b", "u", 2L,
    "d", "W", 1L)
Y
```



# Joins

Joins give us the ability to combine and filter multiple datasets on common column(s). 

## `starwarsdb` 

The code below creates and displays the data model for the tables in this package 

```{r StarWarsDataModel, message=FALSE, warning=FALSE}
StarWarsDataModel <- dm::dm_draw(dm = starwars_dm(), 
                                 view_type = "all",
                                 graph_name = "StarWarsDataModel")
StarWarsDataModel
```

## Create `StarWarsFilms`

Load the `starwarsdb::films` dataset, removing the `opening_crawl` column, and filter to only episodes 4-6. Store these data in `StarWarsFilms`

```{r films-sol}
StarWarsFilms <- starwarsdb::films %>% 
                    select(-c(opening_crawl, director, producer)) %>% 
                    filter(episode_id >= 4 & episode_id <= 6)
StarWarsFilms
```


## `left_join` 

**A `left_join` keeps all of `x`, and joins it to all matching rows from dataset `y`**

Left-join `StarWarsFilms` to `starwarsdb::films_vehicles` on `"title"`, then `select()` the `vehicle` column, then `everything()` else. 

Assign to `StarWarsFilmVeh`

**Example**

```{r example-left-join}
dplyr::left_join(
  x = X, y = Y, 
  by = c("A", "B"))
```


**Solution**

```{r StarWarsFilmsPeople}
dplyr::left_join(x = StarWarsFilms, 
          y = starwarsdb::films_vehicles, 
          by = "title") %>% 
  dplyr::select(vehicle, 
         everything()) -> StarWarsFilmVeh
dplyr::glimpse(StarWarsFilmVeh)
```

We now have a table of vehicles from the episodes 4-6.

## `inner_join`

**The `inner_join()` joins dataset `x` and `y`, and keeps only matching rows from both.**

Use an `inner_join` to join `StarWarsFilmVeh` to `starwarsdb::pilots` on `"vehicle"` and assign it to `StarWarsFilmVehPil`.

**Example**

```{r example-inner-join}
dplyr::inner_join(
  x = X, y = Y, 
  by = c("A", "B"))
```

**Solution**

```{r StarWarsFilmVehPil-sol}
dplyr::inner_join(x = StarWarsFilmVeh, 
           y = starwarsdb::pilots, 
           by = "vehicle") -> StarWarsFilmVehPil
dplyr::glimpse(StarWarsFilmVehPil)
```

Notice the number of rows went from 47 to 39 (because not all vehicles had a pilot in the `starwarsdb::pilots` table)

## `right_join` 

**The `right_join()` includes all the rows in dataset `y`, and joins it the matching rows in dataset `x`**

**Example**

```{r example-right-join}
dplyr::right_join(
  x = X, y = Y, 
  by = c("A", "B"))
```

Use a `right_join()` to join `StarWarsFilmVehPlt` to `starwarsdb::films_planets` on `"title"`, keeping all the rows in `starwarsdb::films_planets`. 

Use `select()` to rename `planet` to `film_planet` and assign this to `StarWarsFilmVehPltPlnt` 

**Solution**

```{r right_join-sol}
dplyr::right_join(x = StarWarsFilmVehPil, 
           y = starwarsdb::films_planets, 
           by = "title") %>% 
  dplyr::select(vehicle:release_date,
         film_planet = planet, 
         pilot) -> StarWarsFilmVehPltPlnt
dplyr::glimpse(StarWarsFilmVehPltPlnt)
```

## `anti_join` 

**The `anti_join()` keeps all rows in dataset `x` that do not have a match in dataset `y`**


Note the last join created empty values for the episodes that weren't in our original `StarWarsFilms` table: 

```{r review-StarWarsFilms}
StarWarsFilms
```

These include: `"The Phantom Menace"`, `"Attack of the Clones"`, and `"Revenge of the Sith"`

```{r filter-is.na}
StarWarsFilmVehPltPlnt %>% 
  filter(title %in% c("The Phantom Menace",
    "Attack of the Clones", 
    "Revenge of the Sith"))
```


Use the `anti_join()` to join `StarWarsFilmVehPltPlnt` to `StarWarsFilms` on `"title"`, `"episode_id"`, and `"release_date"` to create a dataset with only `title` and `film_planet` 

**Example**

```{r example-anti-join}
dplyr::anti_join(
  x = X, y = Y, 
  by = c("A", "B"))
```

**Solution**

```{r anti_join-sol}
anti_join(x = StarWarsFilmVehPltPlnt, 
            y = StarWarsFilms, 
            by = c("title", "episode_id", "release_date")) %>% 
  select(title, episode_id, film_planet)
```

*these should only contain the `title`s not in `StarWarsFilms`*

