---
title: "Case Study (Part 4)"
comment: "Don't Mess with Texas Part 3: manipulating images with `magick`"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
library(here)
library(janitor)
library(socviz)
library(ggrepel)
library(covdata)
library(showtext)
library(hrbrthemes)
library(rtweet)
# create data folder
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.path = "../img/",
                      tidy = FALSE)
# set width
options(width = 60, max.print = 300)
```


## Texas death row executed offenders website

This continues with the [Texas Department of Criminal Justice data](http://www.tdcj.state.tx.us/death_row/dr_executed_offenders.html), which keeps records of every inmate executed. 

## Packages 

```{r packages, message=FALSE, warning=FALSE}
library(knitr)
library(rmdformats)
library(hrbrthemes)
library(tidyverse)
library(rvest)
library(XML)
library(magrittr)
library(xml2)
library(here)
library(magick)
```


## Import the data

We will load previous .csv file of all executions. 

```{r DirProcessed}
fs::dir_tree("../data/wk10-dont-mess-with-texas/")
```

The code below will import the most recent data.

```{r create-ExecOffenders}
# fs::dir_ls("data/processed/2021-10-25")
ExecOffenders <- readr::read_csv("https://bit.ly/2Z7pKTI")
ExOffndrsComplete <- readr::read_csv("https://bit.ly/3oLZdEm")
```

Wrangle these date variables, 

```{r date-vars}
ExecOffenders <- ExecOffenders %>%
  dplyr::mutate(
    date = lubridate::mdy(date),
    year = lubridate::year(date),
    yday = lubridate::yday(date),
    month = lubridate::month(date, label = TRUE)) 
```


```{r skimr-ExecOffenders}
ExecOffenders %>% skimr::skim()
```

These data are already pretty clean, but we will be using the .jpgs I've downloaded in the `02-iterate-with-download.Rmd` file. 

## The `magik` package

I will be using the [magik](https://cran.r-project.org/web/packages/magick/vignettes/intro.html) package for processing and manipulating these images. I advise checking out the entire vignette for more examples. 

## Create a test image

I picked an offender that was typical, meaning they represented the 'average' person from this sample, and stored it as `test_image` below:

The executed offender is [Stevn Coen Renfro](https://murderpedia.org/male.R/r1/renfro-steven.htm) from Harrison Texas. He was executed on Februrary 9th, 1998.

```{r permanent-test_image, echo=FALSE, include=FALSE}
"http://www.tdcj.state.tx.us/death_row/dr_info/renfrosteven.jpg" -> test_image
```
 
I create `test_magick_img` from `magick::image_read()`, and then go on making the transformations as necessary. 


```{r test_magick_img}
test_magick_img <- magick::image_read(test_image)
test_magick_img
```

This images comes up in the viewer pane.

*TIP: come up with a naming convention for each step so you can use RStudio's viewer pane to see the manipulations.*

## Basic transformations

These functions are for basic image movement/manipulations you would do with any basic photo editing app.

### Crop with `magick::image_crop()`

Now I want to remove the text and focus on the mugshot. This might need to be adjusted slightly for each new `test_magick_img`.

```{r test_magick_crop_750_x_1000_10}
# crop this image
test_magick_crop_750_x_1000_10 <- magick::image_crop(
  image = test_magick_img,
  geometry = "750x1000+10"
)
test_magick_crop_750_x_1000_10
```

This should have trimmed the extra space off the bottom of the image. 

### Rotate with `magick::image_rotate()`

I want to rotate this image by 90 degrees. 

```{r test_magick_rotate90}
# rotate this image
test_magick_rotate90 <- magick::image_rotate(test_magick_crop_750_x_1000_10,
  degrees = 90
)
test_magick_rotate90
```

Now I want to remove the rest of the text and focus on the mugshot. This might need to be adjusted slightly for each new `test_image`.

```{r test_magick_crop_850_x_950_450}
# crop this image
test_magick_crop_850_x_950_450 <- magick::image_crop(
  image = test_magick_rotate90,
  geometry = "850x950+450"
)
test_magick_crop_850_x_950_450
```



Now I will rotate this image back to center (`image_rotate` again) and flip it using `magick::image_flip()`

```{r test_magick_rotate270}
# rotate this image
test_magick_rotate270 <- magick::image_rotate(test_magick_crop_850_x_950_450,
  degrees = 270)
# rotate this image
test_magick_flip <- magick::image_flip(test_magick_rotate270)
test_magick_flip
```




I'll crop the rest of the text out of the image, and trim the whitespace for the plot. 

```{r test_magick_crop_750_x_200_10}
# crop this image
test_magick_crop_750_x_200_10 <- magick::image_crop(
  image = test_magick_flip,
  geometry = "750x200+10"
)
test_magick_crop_750_x_200_10
```

Flip this image again.

```{r test_magick_flip2}
# flip this image again
test_magick_flip2 <- magick::image_flip(test_magick_crop_750_x_200_10)
test_magick_flip2
```

Rotate another 270:

```{r test_magick_rotate270v2}
# rotate to remove the dot
test_magick_rotate270v2 <- magick::image_rotate(test_magick_flip2,
  degrees = 270
)
test_magick_rotate270v2
```


Now crop out the last little bit of the document. 

```{r test_magick_crop_640_x_352_10}
# crop the dot out
test_magick_crop_640_x_352_10 <- magick::image_crop(
  image = test_magick_rotate270v2,
  geometry = "650x352+10"
)
test_magick_crop_640_x_352_10
```


Rotate back to center 

```{r test_magick_rotate90v02}
# rotate back to center
test_magick_rotate90v02 <- magick::image_rotate(test_magick_crop_640_x_352_10,
  degrees = 90
)
test_magick_rotate90v02
```

### Image effects with `magick`

Now we will use `magick::image_trim()` to clean the image up a bit. 

```{r test_magick_clean}
# Here we will trim the image up a bit with the `fuzz` argument
test_magick_clean <- magick::image_trim(
  image = test_magick_rotate90v02,
  fuzz = 1
)
test_magick_clean
```


Now that I have all the trimming on and cropping done, I will add some effects for the `ggplot2` image. 

I want the image to be a bit more subdued, so I will use `magick::image_modulate()` and `magick::image_flatten()` to create these effects. 

```{r test_magick_final}
test_image_modulate <- magick::image_modulate(test_magick_clean,
  brightness = 100,
  saturation = 25,
  hue = 20
)
# test_image_modulate
test_magick_final <- magick::image_flatten(test_image_modulate,
  operator = "Threshold"
)
test_magick_final
```


Export image:

```{r export-test_magick_final}
magick::image_write(image = test_magick_final, 
                    path = "../img/test_magick_final.png")
```

## Data for plot

I want to graph the number of executions over time (year) by race. I can do this by getting a grouped data from using `dplyr`'s functions.

## Plot executions over time

I create `base_ggplot2` as the basic plot I want as a layer for the image to appear on top of.

```{r PlotExecOffender}
# Scatter plot
# colors_brewer
PlotExecOffender <- ExecOffenders %>%
  # remove 'Other'
  filter(race != "Other") %>% 
  dplyr::group_by(race, year) %>% 
      dplyr::summarise(
            ex_x_race_year = sum(n())) %>% 
    dplyr::arrange(desc(ex_x_race_year)) %>% 
  ungroup() 
head(PlotExecOffender)
```

## Start with labels! 

Build the labels first: 

```{r labs_executed_offndrss}
labs_executed_offndrss <- ggplot2::labs(
    title = "Texas Justice",
    subtitle = "Executions (1980-2020) in Texas",
    caption = "source: http://www.tdcj.state.tx.us/death_row/index.html",
    x = NULL,
    y = "Executions"
    )
```

## Build a base plot 

Now we can start with a base plot:

```{r create-base_ggplot2, fig.width=7, fig.height=5}
base_ggplot2 <- PlotExecOffender %>% 
  ggplot2::ggplot(data = ., aes(
    y = ex_x_race_year,
    x = year,
    color = race
  )) + 
  labs_executed_offndrss
base_ggplot2
```

### Adding data to canvas

We can now add the data to the canvas. 

Color palettes: http://applied-r.com/rcolorbrewer-palettes/

```{r gg_executions_year_01, fig.width=7, fig.height=5}
library(RColorBrewer)
gg_executions_year_01 <- base_ggplot2 + 
  # add the lines
  ggplot2::geom_line(aes(color = race), size = 2) + 
  # add the points
  ggplot2::geom_point(aes(color = race), 
                      size = 1.2,
                      alpha = 1/3) +
  # set the axes
  ggplot2::scale_x_continuous(breaks = seq(1982, 2020, 4)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 22, 2)) +
  # add the themes
  ggplot2::theme(
    legend.position = "bottom",
    legend.direction = "horizontal",
    legend.title = element_blank(),
    panel.grid.major = element_blank()
    # panel.grid.minor = element_blank()
  ) + 
  scale_color_brewer(palette = "RdBu")
gg_executions_year_01
```

This looks OK, but the white line is hard to see, so we will try another color palette in the example below:

### Example: overplot using `ggpubr` package

The first example I'll plot will use image as the 'canvas'. This requires exporting the image as a .jpeg, then reloading it and using the [`ggpubr`](http://www.sthda.com/english/articles/24-ggpubr-publication-ready-plots/) package. 

```{r image_write, eval=TRUE, echo=TRUE}
library(jpeg)
# 1) export the `test_magick_final` file,
magick::image_write(test_magick_final,
  path =
    paste0("../img/",
      base::noquote(lubridate::today()),
      "-test_magick_final",
      format = ".jpg"
    )
)
# 2) then read it back in as an `jpeg::readJPEG()`.
fs::dir_ls("../img", regexp = lubridate::today())
```

```{r readJPEG-test_magick_final_file}
test_magick_final_file <- list.files("../img", 
                                     pattern = as.character(lubridate::today()), 
                                     full.names = TRUE)
# test_magick_final_file
imgJPEG <- jpeg::readJPEG(test_magick_final_file)
```

### Build new labels 

We want to adjust our labels to include the link about the offender in the background. 

```{r labs_executed_offndrss_02}
labs_executed_offndrss_02 <- ggplot2::labs(
    title = "Texas Justice (executions from 1980-2020)",
    subtitle = "source: http://www.tdcj.state.tx.us/death_row/index.html",
    caption = "Stevn Coen Renfro: https://murderpedia.org/male.R/r1/renfro-steven.htm",
    x = NULL,
    y = "Executions"
    ) 
```

### Add a background image

Now I can add the `imgJPEG` after the base layer (but before I map the `geom_line()` and `geom_theme()`) using `ggpubr::background_image()`. 

```{r gg_executions_year_02, fig.width=7, fig.height=5}
library(hrbrthemes)
library(ggpubr)
gg_executions_year_02 <- base_ggplot2 +
  # this is the image for the background
  ggpubr::background_image(imgJPEG) +

  # add the lines
  ggplot2::geom_line(aes(color = race), alpha = 3/4, 
                     size = 0.8) + 
  # add the points
  ggplot2::geom_point(aes(color = race), 
                      size = 1.3,
                      alpha = 1/3) +
  # set the axes
  ggplot2::scale_x_continuous(breaks = seq(1982, 2020, 4)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 22, 2)) +
  # add the themes
  hrbrthemes::theme_ft_rc() + 
  # add the labels
  labs_executed_offndrss_02 
gg_executions_year_02
```


### Add manual color scale

We will add a manual color scale to this plot with `scale_colour_discrete()`. 

```{r gg_executions_year_03, fig.width=7, fig.height=5}
gg_executions_year_03 <- gg_executions_year_02 + 
  # add a manual color scale
  scale_colour_discrete(type = c("red", "blue", "black"))
gg_executions_year_03
```

```{r gg_executions_year_03.png, echo=FALSE, include=FALSE, results='hide'}
ggsave(
  filename = "../img/gg_executions_year_03.png",
  width = 9,
  height = 7,
  units = "in",
  device = "png"
)
```

### Example: Add labels using `ggrepel`

We also want to label the specific point on the graph where Steven Renfro was executed. We can do this by filtering the `ExecOffenders` to only his row in that dataset, then creating a `label` variable with `paste0()`.

```{r label}
ExecOffLabel <- ExecOffenders %>% 
  filter(last_name == "Renfro") %>%
  mutate(label = paste0(last_name, ", ", first_name, ": executed on ", date)) 
ExecOffLabel
```

We now want to limit the columns in `ExecOffLabel` to *only those values* on the graph, which are `last_name`, `first_name`, `race`, `year`, `label`

```{r ExecOffLabel}
ExecOffLabel <- ExecOffLabel %>% 
  select(last_name, first_name, race, year, label)
ExecOffLabel
```

But wait, we grouped the `ExecOffenders` by race and year to create `PlotExecOffender` (review this code below)

```{r review}
PlotExecOffender <- ExecOffenders %>%
  # remove 'Other'
  filter(race != "Other") %>% 
  dplyr::group_by(race, year) %>% 
      dplyr::summarise(
            ex_x_race_year = sum(n())) %>% 
    dplyr::arrange(desc(ex_x_race_year)) %>% 
  ungroup() 
head(PlotExecOffender)
```

So in order to get that `ex_x_race_year` variable, we will `left_join()` the columns in `ExecOffLabel` to the columns in `PlotExecOffender`. Fortunately, `dplyr` is smart enough to know what we want to join on.

```{r ExecOffLabelData}
ExecOffLabelData <- left_join(x = ExecOffLabel, y = PlotExecOffender)
ExecOffLabelData
```

We see the message: `Joining, by = c("race", "year")` and we now have all the variables we need to add the labels to the graph. Now we load the `ggrepel` package and use the `ggrepel::geom_label_repel()` layer, with `data` set to `ExecOffLabelData`. 

We also add *another* `geom_point()` layer with `ExecOffLabelData` and increase the `size` of the dot so it stands out. 

```{r gg_executions_year_04, fig.width=7, fig.height=5}
library(ggrepel)
gg_executions_year_04 <- base_ggplot2 +
  # this is the image for the background
  ggpubr::background_image(imgJPEG) +

  # add the lines
  ggplot2::geom_line(aes(color = race), alpha = 1/2, 
                     size = 1.5) + 
  # add the points
  ggplot2::geom_point(aes(color = race), 
                      size = 1.2,
                      alpha = 1/4) +
     ggplot2::geom_point(data = ExecOffLabelData,
                     aes(color = race), 
                      size = 10,
                      alpha = 1/4,
                     show.legend = FALSE) +
     # add value label
     geom_label_repel(data = ExecOffLabelData,
              aes(x = year, y = ex_x_race_year,
                  label = label),
                  # set color and size...
                  color = "black",
                  size = 2) +
  # set the axes
  ggplot2::scale_x_continuous(breaks = seq(1982, 2020, 4)) +
  ggplot2::scale_y_continuous(breaks = seq(0, 22, 2)) +
  # add the themes
  hrbrthemes::theme_ft_rc(axis_text_size = 10, axis_title_size = 12) + 
  # add color scale
  scale_colour_discrete(type = c("red", "blue", "black")) +
  # add plot labels
  labs_executed_offndrss_02
gg_executions_year_04
```


```{r gg_executions_year_04.png, echo=FALSE, include=FALSE, results='hide'}
ggsave(
  filename = "../img/gg_executions_year_04.png",
  width = 9,
  height = 7,
  units = "in",
  device = "png"
)
```
