---
title: "Data Visualization with ggplot2 (part 3)"
comment: "*advanced facetting and small multiples*"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---


```{r , include=FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
library(here)
library(janitor)
library(socviz)
library(ggrepel)
library(covdata)
library(showtext)
library(hrbrthemes)
# create data folder
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.path = "img/",
                      tidy = FALSE)
# set width
options(width = 60, max.print = 60)
```

# Outline

1. Import Apple Mobility Data 
2. Facets  
3. Complete themes  
4. Adjusting colors   
5. `geofacet`  

# Materials 

View the slides for this section [here](https://mjfrigaard.github.io/csuc-data-journalism/slides.html).

# Import `TidyApple` data 

We're going to import the data from the previous exercises.  

```{r import-TidyApple-TopUSCities, message=FALSE, warning=FALSE}
TidyApple <- vroom::vroom("https://bit.ly/3IiwcJs", delim = ",")
TopUSCities <- vroom::vroom("https://bit.ly/3d1zEd0", delim = ",")
```


# Advanced Facets

In the previous lesson, we introduced the `facet_wrap()` function for viewing the relationship between two variables across the levels of a categorical variable. In the next section, we're going to show how faceting can be used to explore 'small multiples' in a dataset with variation across multiple levels.

## `facet_wrap()` {.tabset}

Now that we have a graph we can use to compare the `8` cities, we will use `facet_wrap` to create a subplot for each level of `region`.

### exercise

Fill in the `facet_wrap()` (note the use of the `~`) function with `region` and set the `ncol` to `2`.

```{r facet_wrap-ex, eval=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(~ _______,  ncol = _) +
  lab_top_cities
```

### solution

```{r lab_top_cities-sol}
lab_top_cities <- labs(x = "Date", y  = "Direction Requests",
     title = "Trends of Relative Activity in Selected US Cities",
     subtitle = "NY, LA, CH, HOU, PHA, PHL, SATX, SD",
     color = "Type")
```

```{r facet_wrap-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
TopUSCities %>% 
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(~ region,  ncol = 2) +
  lab_top_cities
```

### exercise

Now map *both* `region` and `trans_type` to `facet_wrap()` and set the `ncol` to `6`.

```{r facet_wrap-2-ex, eval=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(_______ ~ _______, ncol = _) +
  lab_top_cities
```


### solution

```{r facet_wrap-2-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(region ~ trans_type, ncol = 6) +
  lab_top_cities
```


## Complete Themes  {.tabset}

`ggplot2` comes with a few default theme settings. Read more about them [here](https://ggplot2.tidyverse.org/reference/ggtheme.html). These themes come with arguments to adjust the `font_size` and `font_family` on our plot.

```{r theme_minimal-settings, eval=FALSE}
theme_minimal(
  base_size = 11,
  base_family = "",
  base_line_size = base_size/22,
  base_rect_size = base_size/22
)
```


### exercise

We'll use the `theme_minimal()` function to reduce the number of `elements` on our graph (don't add any arguments).

```{r theme_minimal-ex, eval=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(region ~ trans_type, ncol = 6) +
  ______________() +
  lab_top_cities
```

### solution

```{r theme_minimal-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(region ~ trans_type, ncol = 6) +
  theme_minimal() +
  lab_top_cities
```


## Theme elements {.tabset}

Themes (accessible with the `theme` function) give up the ability to customize various `element_()` settings, which affect the 'look' of our graph.

### exercise

We will start by moving the `legend.position` to `"top"` so it's not crowding the `x` axis on our graph.

```{r theme-ex, eval=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(region ~ trans_type, ncol = 6) +
  theme_minimal() +
  theme(___________ = _____) +
  lab_top_cities
```

### solution

```{r theme-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
TopUSCities %>%
  ggplot(aes(x = date, y = dir_request,
             group = region,
             color = trans_type)) +
  geom_line(size = 0.1) +
  geom_hline(yintercept = 100, size = 0.2, color = "gray20") +
  facet_wrap(region ~ trans_type, ncol = 6) +
  theme_minimal() +
  theme(legend.position = "top") +
  lab_top_cities
```

Notice we pass the `theme()` layer after the `theme_minimal()` layer (if we ordered these the other way around, `theme_minimal()` would overwrite any custom settings we made in `theme()`).

## Small Multiples  {.tabset}

What if we wanted to facet more than 8 cities? Fortunately, we have a the [`ggforce`](https://ggforce.data-imaginist.com) and  [`geofacet`](https://hafen.github.io/geofacet/index.html) packages for doing just that!

```{r packages-multiple-facets, message=FALSE, warning=FALSE}
library(sf)
library(geofacet)
library(ggforce)
library(jcolors)
```

**Building the Graph Data**

1. We will start by filtering the `TidyApple` to only the 50 US states (we've removed three US territories) and storing these data in `TidyAppleUS`.

2. Next we limit the date range to the beginning of the shelter in place (from February 1, 2020 to May 1, 2020). These data get stored in `TidyAppleUST1`.

3. We then create a dataset with only `"transit"` direction requests, and we count these by state (`sub_region`), arrange the data descending with `sort = TRUE`, and take the top 25 rows (`Top25TransitStates`).

Finally, we filter `TidyAppleUST1` using the 25 states in `Top25TransitStates` to create our graphing dataset, `Top25TransitUSAllT1`.

```{r TidyAppleUST1-skimr, message=FALSE, warning=FALSE}
# create only US states (TidyAppleUS)
TidyApple %>%
  filter(country == "United States" &
         !sub_region %in% c("Guam", "Puerto Rico",
                            "Virgin Islands")) -> TidyAppleUS
# create shelter in place time 1 (TidyAppleUST1)
TidyAppleUS %>%
  filter(date >= as_date("2020-02-01") &
           date <= as_date("2020-05-01")) -> TidyAppleUST1

# create top 25 states (Top25TransitStates)
Top25TransitStates <- TidyAppleUST1 %>%
  filter(trans_type == "transit") %>%
  count(sub_region, trans_type, sort = TRUE) %>%
  head(25)

# filter T1 to states with the most transit requests (Top25TransitUSAllT1)
TidyAppleUST1 %>%
  filter(sub_region %in%
           unique(Top25TransitStates$sub_region)) -> Top25TransitUSAllT1

Top25TransitUSAllT1 %>% skimr::skim()
```

### exercise

- set `title` to `"States With Highest Transit Direction Requests"`

- set `subtitle` to `"Top 25 states based on number of total transit requests"`

```{r lab_facet_wrap_paginate-ex, eval=FALSE}
lab_facet_wrap_paginate <- labs(
       x = "Date", y = "Direction Requests",
       title = _____________________________________,
       subtitle = _____________________________________)
```

### solution

```{r lab_facet_wrap_paginate}
lab_facet_wrap_paginate <- labs(
       x = "Date", y = "Direction Requests",
       title = "States With Highest Transit Direction Requests",
       subtitle = "Top 25 states based on number of total transit requests")
```

### exercise


Inside `ggforce::facet_wrap_paginate()`:

- map `sub_region` as the variable to facet using the `~`

- map `5` to `ncol`

- map `"free_y"` to `scales`

Inside `theme()`

- map `element_blank()` to `panel.border` and `panel.background`

- map `element_text(size = 6)` to `axis.text.x` and `axis.text.y`

- map `element_text(colour = 'black')` to `strip.text`

- map `element_rect(fill = "gray93")` to `strip.background`

- map `"top"` to `legend.position`


```{r facet_wrap_paginate-ex, eval=FALSE}
Top25TransitUSAllT1 %>%
  # global settings
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  # lines
  geom_hline(yintercept = 100, size = 0.3, color = "black") +
  geom_line(size = 0.2) +
  # faceting
  ggforce::facet_wrap_paginate(~ __________,
                               ncol = _,
                               scales = _______) +
  # theme settings
  theme(__________ = __________(),
        __________ = __________(),

        __________ = __________(size = _),
        __________ = __________(size = _),

        __________ = __________(colour = __________),

        __________ = __________(fill = __________),

        __________ = __________) +
  # labels
  lab_facet_wrap_paginate
```

### solution

```{r facet_wrap_paginate-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
Top25TransitUSAllT1 %>%
  # global settings
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  # lines
  geom_hline(yintercept = 100, size = 0.3, color = "black") +
  geom_line(size = 0.2) +
  # faceting
  ggforce::facet_wrap_paginate(~ sub_region,
                               ncol = 5,
                               scales = "free_y") +
  # theme settings
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        strip.text = element_text(colour = 'black'),
        strip.background = element_rect(fill = "gray93"),
        legend.position = "top") +
  # labels
  lab_facet_wrap_paginate
```

## Adjusting Colors {.tabset}

Changing the colors on graphs gives us the ability to further customize their look. We can set these manually, or use one of the many complete color palettes from a user-written package. Below we'll use the [`jcolors` package](https://jaredhuling.org/jcolors/index.html) to highlight the `transit` direction requests from the previous graph.

### exercise

- add `scale_color_jcolors()` and play with the `palette` argument to make the graph look like the `solution`.

```{r scale_color_jcolors-ex, eval=FALSE}
Top25TransitUSAllT1 %>%
  # global settings
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  # lines
  geom_hline(yintercept = 100, size = 0.3, color = "black") +
  geom_line(size = 0.2) +
  # faceting
  ggforce::facet_wrap_paginate(~ sub_region,
                               ncol = 5,
                               scales = "free_y") +
  # theme settings
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        strip.text = element_text(colour = 'black'),
        strip.background = element_rect(fill = "gray93"),
        legend.position = "top") +
  # adjust colors
  __________________________(palette = ____) +

  lab_facet_wrap_paginate
```

### solution

```{r scale_color_jcolors-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
Top25TransitUSAllT1 %>%
  # global settings
  ggplot(aes(x = date, y = dir_request,
             group = trans_type,
             color = trans_type)) +
  # lines
  geom_hline(yintercept = 100, size = 0.3, color = "black") +
  geom_line(size = 0.2) +
  # faceting
  ggforce::facet_wrap_paginate(~ sub_region,
                               ncol = 5,
                               scales = "free_y") +
  # theme settings
  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        strip.text = element_text(colour = 'black'),
        strip.background = element_rect(fill = "gray93"),
        legend.position = "top") +

  scale_color_jcolors(palette = "pal3") +

  lab_facet_wrap_paginate
```

## Extra: `geofacet` {.tabset}

We're now going to look at all 50 states using the `facet_geo()` function from the `geofacet` package. To make this graph easier to interpret, we're going to focus only on walking mobility data, and adjust the `dir_request` value to absolute change from baseline (set to `100` on `2020-01-13`).

### Adjusted Walking Data

Create the `USWalkingAdj` data by filtering the `trans_type` to `"walking"` and creating two new variables: `above_below` (a logical indicator for values being above or below the baseline value of `100`), and `dir_request_adj` (the adjusted direction request value).

### exercise

Fill in the correct variables in the wrangling steps below:

```{r USWalkingAdj-ex, eval=FALSE}
USWalkingAdj <- TidyAppleUS %>%
  filter(trans_type == _________) %>%
  mutate(above_below = _________ < 100,
         dir_request_adj = _________ - 100)

USWalkingAdj %>%
  skimr::skim()
```

### solution

```{r USWalkingAdj-sol}
USWalkingAdj <- TidyAppleUS %>%
  filter(trans_type == "walking") %>%
  mutate(above_below = dir_request < 100,
         dir_request_adj = dir_request - 100)

USWalkingAdj %>%
  skimr::skim()
```

### exercise

Assign the following to the labels:

- set `"US Trends in Mobility Data for Walking (Adjusted)"` to `title`

- set `"https://covid19.apple.com/mobility"` to `caption`

```{r , lab_facet_geo-ex, eval=FALSE}
lab_facet_geo <- labs(x = "Date",
                      y = "Direction Requests (Walking)",
       title = ____________________________________________,
       subtitle = paste0("Original data uses 100 as baseline for usage at ",
                         min(USWalkingAdj$date)),
       caption = ____________________________________________,
       fill = "Below Baseline")
```

### solution

```{r lab_facet_geo-sol}
lab_facet_geo <- labs(x = "Date",
                      y = "Direction Requests (Walking)",
       title = "US Trends in Mobility Data for Walking (Adjusted)",
       subtitle = paste0("Original data uses 100 as baseline for usage at ",
                         min(USWalkingAdj$date)),
       caption = "https://covid19.apple.com/mobility",
       fill = "Below Baseline")
```

### exercise

- set the colors in `color_bl_or` as `c("#8470FF", "#7FFFD4")`

- set `yintercept` to `0` in `geom_hline()`

- set the `values` in `scale_fill_manual()` to `color_bl_or`

- map `sub_region` to `facet_geo` using `~`

Inside `theme()`

- set the `panel.border` and `panel.background` to `element_blank()`
- set the `axis.text.x` and `axis.text.y` to `element_text(size = 6)`
- set the `strip.text.x` to `element_text(size = 7)`
- set `strip.text` to `element_text(colour = 'white')`
- set `strip.background` to `element_rect(fill = "black")`
- set `legend.position` to `"bottom"`

```{r facet_geo-ex, eval=FALSE}
# set colors
color_bl_or <- c(____________, ____________)

USWalkingAdj %>%
  ggplot(aes(x = date, y = dir_request_adj,
             group = sub_region, fill = above_below)) +

  geom_col() +

  geom_hline(yintercept = _, color = "gray7") +

  scale_fill_manual(values = ____________) +

  facet_geo(~ sub_region) +

  theme_bw() +

  theme(______________ = ______________(),
        ______________ = ______________(),

        ______________ = ______________(size = _),
        ______________ = ______________(size = _),
        ______________ = ______________(size = _),

        ______________ = ______________(colour = ______________),
        ______________ = ______________(fill = ______________),

        ______________ = ______________) +

  lab_facet_geo
```

### solution

```{r facet_geo-sol, fig.height=8, fig.width=10, out.height='100%', out.width='100%', message=FALSE, warning=FALSE}
# set colors
color_bl_or <- c("#8470FF", "#7FFFD4")

USWalkingAdj %>%
  ggplot(aes(x = date, y = dir_request_adj,
             group = sub_region, fill = above_below)) +

  geom_col() +

  geom_hline(yintercept = 0,
             color = "gray7") +

  scale_fill_manual(values = color_bl_or) +

  facet_geo(~ sub_region) +

  theme_bw() +

  theme(panel.border = element_blank(),
        panel.background = element_blank(),
        axis.text.x = element_text(size = 6),
        axis.text.y = element_text(size = 6),
        strip.text.x = element_text(size = 7),
        strip.text = element_text(colour = 'white'),
        strip.background = element_rect(fill = "black"),
        legend.position = "bottom") +

  lab_facet_geo
```

# Wrap Up

**Original Question:** *How has COVID changed our modes of transportation?*

1. Which graphs do you feel are best at answering this question? Why?

2. What other information (tables, annotations, etc.) would you include with the graphs?


