---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Week 3: R Markdown in Practice"
author: "<Your Name>"
output: html_document
---

```{r , include=FALSE}
library(tidyverse)
library(ipcwswitch)
library(skimr)
library(listviewer)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")
```

We'll be using R Markdown for most of the assignments in this course. As you'll discover, R Markdown is a powerful tool because it can combine prose/text, R code chunks and outputs (like figures and tables).

In this assignment, you will get a broad overview of what R markdown can do. You can review the slides for this section [here](https://mjfrigaard.github.io/FA22-JOUR301/slides/intro-to-rmarkdown.html#1), 

When you've reached the end of this document, click on the "Knit" icon in R Studio and review the output.

**Edit only <Your Name> above and below this line.**

---------------------------------------------

# Project setup 

Follow the instructions on using `goodenuffR` on the course website [here]() or on the [package website.](https://mjfrigaard.github.io/goodenuffR/). 

Create a new R Markdown file and name it with the following conventions: 

`YYYY-MM-DD-LASTNAME-wk3-rmarkdown-assignment.Rmd`

*Replace the date (`YYYY-MM-DD`) and enter your last name (`LASTNAME`).*

Save this file in the top-level folder of the project: 

```{r tree, echo=TRUE, eval=FALSE}
├── CITATION
├── LICENSE
├── README.Rmd
├── YYYY-MM-DD-LASTNAME-wk3-rmarkdown-assignment.Rmd <- YOUR HW FILE!
├── code
│   ├── 01-import.R
│   ├── 02-tidy.R
│   ├── 03-wrangle.R
│   ├── 04-visualize.R
│   ├── 05-model.R
│   ├── 06-communicate.R
│   └── runall.R
├── data
│   ├── README.md
│   └── raw
├── docs
│   ├── changelog.txt
│   ├── manuscript.Rmd
│   └── notebook.Rmd
├── requirements.txt
├── results
│   ├── figures
│   ├── manuscript
│   └── tables
└── wk_3assignment.Rproj
```

# Downloading the files for this assignment 

Download the files for this assignment by running the code chunk below:

```{r download.file, eval=FALSE}
download.file(
  url = "https://github.com/mjfrigaard/FA22-JOUR301/raw/main/data/starwars.rds", 
  destfile = "data/starwars.rds")
```

## Parameters

Add the following code to the YAML header (at the top of this file):

```{.yaml}
params:
  data_dir: !r file.path("data/starwars.rds")
  list_vars: !r c("films", "vehicles", "starships")
```

These parameters will give us global control over the data we will be importing (even if that file changes in the future).

## Importing and Viewing Data 

```{r StarWars, eval=FALSE}
StarWars <- readr::read_rds(file = params$data_dir)
```

Note that we've loaded these data using the parameters we've defined above.

### Help files

Read about the variables in the `StarWars` dataset using `??starwars`

### `glimpse()`

View a basic transposed display of the `StarWars` data with `dplyr`'s `glimpse()` function.

```{r glimpse-StarWars, eval=FALSE}
dplyr::glimpse(StarWars)
```

`glimpse()` shows us the format and first few values of each variable in `StarWars`.

### `skim()`

Below is a `skimr::skim()` view of the `StarWars` data. We can see each variable broken down by type, along with some summary information.

```{r skimr, eval=FALSE}
install.packages("skimr")
library(skimr)
```


```{r skim-StarWars, eval=FALSE}
skimr::skim(StarWars)
```

Read more [here](https://docs.ropensci.org/skimr/).

### `jsonedit()`

If you have JSON or lists (non-rectangular data) in R, sometimes these objects can be hard to visualize. The `jsonedit()` function from `listviewer` makes this easier by giving us an interactive display to click-through.

```{r jsonedit-StarWars, eval=FALSE}
library(listviewer)
listviewer::jsonedit(listdata = StarWars, mode = "view")
```

## Caching data 

Include `cache=TRUE` in `StarWars` code chunk above and re-knit this chunk with the new `cache` option.

### Dataset size

Use `object.size()` from the `utils` package (which is loaded by default).

```{r object.size-StarWars, eval=FALSE}
library(utils)
object.size(StarWars)
```

Another option is using the `inspect_mem()` function from the  [`inspectdf` package](https://alastairrushworth.github.io/inspectdf/).

```{r inspect_mem-StarWars, eval=FALSE}
install.packages("inspectdf")
library(inspectdf)
```

```{r , eval=FALSE}
inspectdf::inspect_mem(df1 = StarWars) %>%
  inspectdf::show_plot(text_labels = TRUE,
                       col_palette = 1)
```

### Get a subset of StarWars

Create a `StarWarsSmall` dataset from the saved the names of the list-columns in `params$list_vars`.

```{r StarWarsSmall, eval=FALSE}
StarWarsSmall <- # enter the code from this section
```

Check the size of the new `StarWarsSmall` data by comparing it to the original `StarWars` dataset.

```{r inspect_mem-StarWars-StarWarsSmall, eval=FALSE}
inspectdf::inspect_mem(df1 = StarWars, df2 = StarWarsSmall) %>%
  inspectdf::show_plot(text_labels = TRUE, col_palette = 1)
```

### Cache path

When we cache data, a new folder named `your-file-name` + `_cache/html/` is created in the same directory as our R Markdown file. 

View the `<file name>_cache/html/` folder contents below:

```{r cache, eval=FALSE}
fs::dir_tree(".")
```

## Figures 

We want to keep track of all the visualizations we create in our report. R Markdown comes with multiple options for controlling the size, location, and quality of images in our reports.

### Size

Below we visualize the average BMI by `species` and `gender` in the Star Wars universe. We also load the [`hrbrthemes` package](https://cinc.rud.is/web/packages/hrbrthemes/) to give us more control over the aesthetics in our plot.

```{r hrbrthemes, eval=FALSE}
install.packages("hrbrthemes")
library(hrbrthemes)
```


Adjust the size of the figure with `fig.height=5.5` and `fig.width=8`. 

```{r gg_avg_bmi_spec_gend, eval=FALSE, fig.height=, fig.width=, out.width='100%', out.height='100%'}
StarWars %>%
  dplyr::filter(!is.na(mass) & !is.na(height) & !is.na(species)) %>%
  dplyr::mutate(bmi = mass / ((height / 100)  ^ 2)) %>%
  dplyr::group_by(species, gender) %>%
  dplyr::summarize(mean_bmi = mean(bmi, na.rm = TRUE)) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(desc(mean_bmi)) %>%
  dplyr::mutate(species = reorder(species, mean_bmi)) %>%
  ggplot2::ggplot(aes(x = mean_bmi, y = species,
                      color = as.factor(species),
                      group = gender)) +
  ggplot2::geom_point(show.legend = FALSE) +
  ggplot2::facet_wrap(. ~ gender, scales = "free") +
  ggplot2::labs(title = "Average BMI in Star Wars Universe",
                subtitle = "Grouped by species and gender",
                caption = "source = https://swapi.dev/",
                x = "Mean BMI", y = "Species") +
  hrbrthemes::theme_ipsum_rc(axis_text_size = 9,
                             axis_title_size = 13,
                             strip_text_size = 13) -> gg_avg_bmi_spec_gend
gg_avg_bmi_spec_gend
```


### Location

Manually specify the figures to be saved in `results/figures` with `fig.path="results/figures/`. If we're setting a folder for the figures, we can do it in the code chunk,

### Interactive figures

Recreate the plotly graph below 

```{r ggplotly, eval=FALSE, fig.height=5, fig.width=7}
library(plotly)

```


Knit this file and view the results. 




