---
#############################################################
#                                                           #
#   Click on "Knit" in RStudio to render this worksheet.    #
#                                                           #
#############################################################
title: "Week 5: ggplot2 part 1"
author: "<Your Name>"
output: html_document
---

```{r , include=FALSE}
library(tidyverse)
library(ipcwswitch)
library(skimr)
library(listviewer)
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE,
                      size = "Huge")
```

In this assignment, we're going to follow the steps in the first set of exercises for ggplot2. You can review the slides for this section [here](https://mjfrigaard.github.io/FA22-JOUR301/slides/ggp2-part-01.html#1), 

In all of the exercises below, the code chunks have been set to `eval=FALSE`. In order to get credit for this assignment, you'll have to set each of them to `eval=TRUE` (or delete the `eval` argument).

When you've reached the end of this document, click on the "Knit" icon in R Studio and review the output.

**Edit only <Your Name> above and below this line.**

---------------------------------------------

# Project setup 

Follow the instructions for setting up your project with `goodenuffR` on the [course website.](https://mjfrigaard.github.io/FA22-JOUR301/exercises/goodenuffR.html) 

# Load the packages

The main packages we're going to use are `dplyr`, `tidyr`, and `ggplot2`. These are all part of the `tidyverse`, so we'll import this package below:

```{r packages, eval=FALSE}
install.packages(c("tidyverse", "janitor", "ggridges"))
library(tidyverse)
library(janitor)
library(ggridges)
```

# Import

Import the data below

```{r import-AppleMobRaw, eval=FALSE}
AppleMobRaw <- readr::read_csv("https://bit.ly/3ALws1X")
```

Place the `AppleMobRaw` dataset in the code below to view the `head()` and `tail()`

```{r head-AppleMobRaw, eval=FALSE}
head(_____)
```

```{r tail-AppleMobRaw, eval=FALSE}
tail(_____)
```

## Tidy the datset

The code chunk below restructures the data into a long (tidy) format. 

```{r TidyApple, eval=FALSE}
TidyApple <- AppleMobRaw %>% 
  tidyr::pivot_longer(cols = -c(geo_type:country), 
                      names_to = "date", values_to = "dir_request") %>% 
    # remove missing country and missing sub-region data
  dplyr::filter(!is.na(country) & !is.na(`sub-region`)) %>% 
  # format date
  mutate(date = lubridate::ymd(date)) %>% 
  # change name of transportation types
  rename(trans_type = transportation_type) %>% 
  # clean names 
  janitor::clean_names() 
```


## Counting

Use `dplyr::count` to count `trans_type` (sort the results)

```{r , eval=FALSE}
TidyApple %>% 
  dplyr::count(_______, sort = _____)
```

# Visualizing Distributions

It's always a good idea to view each variable (or column) in a dataset independently before visualization their relationships. Single variable visualizations show us the 'distribution' of a variables values. 

## Histograms 

Start by creating the labels for the graph. 

```{r lab_hist, eval=FALSE}
lab_hist <- labs(x = "Apple directions requests",
                 y = "Count",
     title = "Distribution of Direction Requests",
     subtitle = "source: https://covid19.apple.com/mobility")
```

Create a histogram of direction requests using `dir_request`

```{r geom_histogram-ex, eval=FALSE}
TidyApple %>% ggplot() + 
  geom_histogram(aes(x = ____________)) + 
  lab_hist
```

## Adjusting Y Axes

Add the `scales::comma` value to the `scale_y_continuous()` function.

```{r scales-ex, eval=FALSE}
library(scales)
TidyApple %>% ggplot() + 
  geom_histogram(aes(x = dir_request)) + 
  scale_y_continuous(labels = __________) +
  lab_hist
```

## Histogram Shape 

Set `bins` to `15`.

```{r bins-15-ex, eval=FALSE}
TidyApple %>% ggplot() + 
  geom_histogram(aes(x = dir_request), bins = __) + 
  scale_y_continuous(labels = scales::comma) +
  lab_hist
```

Set `bins` to `45` and assign it to `gg_hist45`.

```{r bins-45-gg_hist45-ex, eval=FALSE}
TidyApple %>% ggplot() + 
  geom_histogram(aes(x = dir_request), bins = __) + 
  scale_y_continuous(labels = scales::comma) +
  lab_hist -> _____________
```

## Density Plots

Create labels for the distribution of the directions request, filled by the levels of transportation type.

```{r lab_density, eval=FALSE}
lab_density <- labs(x = "Apple directions requests",
                    fill = "Transit Type",
     title = "Distribution of Direction Requests vs. Transportation Type",
     subtitle = "source: https://covid19.apple.com/mobility")
```

Create a density plot of direction requests colored by the type of transportation.

```{r density-ex, eval=FALSE}
TidyApple %>% 
  ggplot() +
  geom_density(aes(x = __________, fill = __________)) + 
  lab_density
```

Adjust the overlapping densities by setting `alpha` to `1/3`. Assign this plot to `gg_density`.

```{r density-alpha-ex, eval=FALSE}
TidyApple %>% 
  ggplot() +
  geom_density(aes(x = dir_request, fill = trans_type), 
               alpha = __________) + 
  
  lab_density -> __________
```

## Ridgeline Plots {.tabset}

Another option is a ridgeline plot (from the `ggridges` package). These display multiple densities.  

```{r lab_ridges, eval=FALSE}
lab_ridges <- labs(
     title = "Direction Requests by Transportation Type",
     subtitle = "source: https://covid19.apple.com/mobility",
     fill = "Transit type",
     x = "Apple directions requests",
     y = "Transportation Types")
```

Set the `fill` argument to Transportation Types and the `alpha` to 1/5.

```{r geom_density_ridges-sol, eval=FALSE, message=FALSE, warning=FALSE, fig.height=4, fig.width=6, out.height='100%', out.width='100%'}
TidyApple %>%  
  ggplot() + 
  geom_density_ridges(aes(x = dir_request, 
                          y = trans_type, 
                          fill = __________), 
                      alpha = _/_) + 
  lab_ridges
```


## Violin Plots 

Another alternative to the density plot is the violin plot.

- assign `"Apple directions requests"` to the `x` axis 
- assign `"Transit Type"` to the `y` axis 

```{r lab_violin-ex, eval=FALSE}
lab_violin <- labs(x = _________________________,
                    y = _________________________,
                   fill = "Transit Type",
     title = "Distribution of Direction Requests vs. Transportation Type",
     subtitle = "source: https://covid19.apple.com/mobility")
```

Add a `geom_violin()` to the code below:

```{r geom_violin-ex, eval=FALSE}
TidyApple %>% 
  ggplot() +
  ____________(aes(y = dir_request, x = trans_type, 
                  fill = trans_type)) + 
  lab_violin
```

The great thing about `ggplot2`s layered syntax, is that we can add `geoms` with similar aesthetics to the same graph! For example, we can see how `geom_violin`s and `geom_boxplot`s are related by adding a `geom_boxplot()` layer to the graph above. 

```{r geom_violin-boxplot-ex, eval=FALSE}
TidyApple %>% 
  ggplot() +
  geom_violin(aes(y = dir_request, x = trans_type, 
                  fill = trans_type), alpha = 1/5) + 
  ___________(aes(y = dir_request, x = trans_type, 
                   color = trans_type)) + 
  lab_violin
```



