---
title: "Case Study (Part 3)"
comment: "Don't Mess with Texas Part 3: Split and wrangle the data"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---

```{r , include=FALSE}
library(tidyverse)
library(skimr)
library(lubridate)
library(here)
library(janitor)
library(socviz)
library(ggrepel)
library(covdata)
library(showtext)
library(hrbrthemes)
library(rtweet)
# create data folder
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      fig.path = "img/",
                      tidy = FALSE)
# set width
options(width = 60, max.print = 300)
```

# Texas death row executed offenders website

This continues with the [Texas Department of Criminal Justice data](http://www.tdcj.state.tx.us/death_row/dr_executed_offenders.html), which keeps records of every inmate executed. 

## The data

These data are imported from the .Rmd we used to scrape the website. These data are in the folder below. 

```{r DirProcessed}
fs::dir_tree("../data/wk10-dont-mess-with-texas/")
```

This will import the most recent data.

```{r create-ExecOffenders}
# fs::dir_ls("data/processed/2021-10-25")
ExecOffenders <- readr::read_csv("https://bit.ly/2Z7pKTI")
ExOffndrsComplete <- readr::read_csv("https://bit.ly/3oLZdEm")
```

## Use `purrr` and `dplyr` to split and export .csv files

This next use of `purrr` and iteration will cover how to:

1. Split the `ExecOffenders` data frame into `ExExOffndrshtml` and `ExExOffndrsjpg` 

2. Save each of these data frames as .csv files

We should have two datasets with the following counts. 

```{r check-jpg_html-counts}
ExecOffenders %>% 
  dplyr::count(jpg_html, sort = TRUE)
```

These are new experimental functions from `dplyr`, and a big shout out to Luis Verde Arregoitia for [his post](https://luisdva.github.io/rstats/export-iteratively/) on a similar topic. 

The `dplyr::group_split()` *"returns a list of tibbles. Each tibble contains the rows of .tbl for the associated group and all the columns, including the grouping variables"*, and I combine it with `purrr::walk()` and `readr::write_csv()` to export each file. 

```{r dplyr-purrr}
ExecOffenders %>% 
  dplyr::group_by(jpg_html) %>% 
  dplyr::group_split() %>% 
  purrr::walk(~.x %>% # we now carry this little .x everywhere we want it 
                      # to go.
                write_csv(path = paste0("../data/", 
                            # processed data folder
                            "wk10-dont-mess-with-texas/processed/",
                            # datestamp
                            base::noquote(lubridate::today()),
                            # folder
                            "/",
                            # datestamp
                            base::noquote(lubridate::today()),
                            # name of file
                            "-ExExOffndrs",
                            # split by this variable
                            base::unique(.x$jpg_html), 
                            # file extension
                            ".csv")))
fs::dir_tree("../data/wk10-dont-mess-with-texas/processed/")
```

### End
