---
title: "Flow Control Statements"
comment: "*change the flow of your code*"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged

always_allow_html: true
---

```{r, echo = FALSE}
# don't name this setup!
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

# Materials 

View the slides for this section [here](https://mjfrigaard.github.io/csuc-data-journalism/slides.html).

View the exercises for this section [here](https://mjfrigaard.github.io/csuc-data-journalism/lessons-exercises.html).

# Flow Control Statements

Flow control functions (`if`, `ifelse`, and `switch`) can be used to control data workflows. Below we'll cover R's common flow control statements: `if()`, `else()`, and `switch()`.

## `if` {.tabset}

The `if()` statement can be used to control exactly what operations are carried out in a given chunk of code.

### `if()` and `cat()`

Below is a straightforward `if()` statement that tests if a given condition is `TRUE`.

```{r if-statement-example-1}
if (TRUE) {
  cat("The if() condition is true")
}
```

What if we wanted to test if a certain vector value (`y`) was greater than `0`?

```{r if-statement-example-2}
x <- 2
if (x > 0) {
  cat("x value is greater than 0")
}
```

### `if()` with `else()`

Now what if we wanted to our code to do something when the if condition was *not* satisfied?

Below we create `y` and set it equal to `5`, then we test to see if `x` is greater than `y`. If it isn't we print the second statement (and the value of `y`)

```{r if-else-statement-example-1}
y <- 5
if (x > y) {
  cat("x is greater than y")
} else {
  cat("y is greater than x, because y = ", y)
}
```

We can extend the `if()` and `else()` conditions as far as we need.  Change the value of `y` to `3` and run two logical tests to see what prints.

```{r if-else-statement-example-2}
y <- 3
if (x > y) {# FALSE
  cat("x is greater than y")
} else if (y > 4) {# FALSE
  cat("y is greater than 4")
} else {# TRUE
  cat("If x is < y, and y is > 4, What is x + y? ", sum(x, y))
}
```

When testing `if()`/`else()` statements, I like to use comments to document what I should expect from each condition.

### vectorized `ifelse()`

Recall that `TRUE` = `1` and `FALSE` = `0`.

```{r test-true-false}
sum(FALSE, FALSE)
sum(TRUE, FALSE)
sum(TRUE, TRUE)
```

We define a vector (`conditions_01`) with multiple values.

```{r conditions_01}
conditions_01 <- c(0, 1, NA, FALSE, TRUE)
conditions_01
```

```{r ifelse-conditions_01}
writeLines(
  ifelse(test = conditions_01,
    yes = "condition is met",
    no = "condition is not met"
  )
)
```

The `ifelse()` condition is tested across *all* the values in `conditions_01` to returns the `yes` or `no` result.

Had we simply used an `if()` and `else()` statement, we get a warning because it's only testing the first element in `conditions_01`.

```{r if-else-condition-1-2}
if (conditions_01) cat(" condition met ") else cat(" condition not met ")
```

This is because the actual condition being tested is `length(conditions_01) > 1`. We can see below this is `FALSE`, so the second result is returned.

```{r if-else-condition-1-length}
# recall frist element in conditions_01
conditions_01[1]
# check length
length(conditions_01[1]) > 1
```

## `switch` {.tabset}

We can use `switch()` in place of nested `if` `else` statements to select among several alternative results.

```r
switch(EXPR = , ... = )
```

The `EXPR` in `switch()` must evaluate to a number or a character vector of length `1`.

### re-write nested `ifelse`

Consider the nested if else statement below:

```{r am_dad_char-am_dad_age}
# define string value
am_dad_char <- "Steven"
am_dad_char
if (am_dad_char == "Stan") {
  am_dad_age <- 42
} else if (am_dad_char == "Francine") {
  am_dad_age <- 40
} else if (am_dad_char == "Steven") {
  am_dad_age <- 15
} else if (am_dad_char == "Hayley") {
  am_dad_age <- 19
} else {
  am_dad_age <- NA
}
# return age
am_dad_age
```

We can re-write this as a `switch()` statement, with the `EXPR` as `am_dad_char`. The alternative results are listed with each character's `age` in `AmericanDad`

```{r switch-am_dad_char}
switch(EXPR = am_dad_char,
  "Stan" = 42,
  "Francine" = 40,
  "Steven" = 15,
  "Haley" = 19
)
```

### `switch()` alternatives

While the `EXPR` needs to be either a string or numeric value, the alternatives (`...`) can be multiple types. See the examples below:

```{r switch-examples}
# switch on a string?
switch("string",
  "string" = "text",
  "logical" = TRUE,
  "number" = 15)
# switch on a logical?
switch("logical",
  "string" = "text",
  "logical" = TRUE,
  "number" = 15)
# switch on a number?
switch("number",
  "string" = "text",
  "logical" = TRUE,
  "number" = 15)
```


