---
title: "Exploratory Data Analysis (EDA) - project template"
author: "Your Name"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    code_folding: show
    theme: journal
    df_print: paged
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(skimr)
# chunk options ----
knitr::opts_chunk$set(warning = FALSE,
                      message = FALSE,
                      eval = FALSE,
                      tidy = FALSE,
                      fig.path = "img/")
# base options ----
base::options(
  tibble.print_max = 78,
  tibble.width = 78,
  scipen = 100000000,
  max.print = 999999
)
```

# Intro 

This R Markdown document template for your exploratory data analysis project. You've selected (or you've been assigned) a dataset, and you're going to use this template to import, wrangle, visualize and communicate the contents of your data. 

The code chunks in this template have been set to `eval=FALSE`. To get them to run, you will need to set them to `eval=TRUE`. 

# Background 

You should have your project set up using the [goodenuffR package](https://github.com/mjfrigaard/goodenuffR). If you haven't, install this package using the code below: 

```{r goodenuffR}
install.packages("devtools")# for install_github()
library(devtools) # load devtools
devtools::install_github("mjfrigaard/goodenuffR") # install goodenuffR
library(goodenuffR) # load goodenuffR
```

Recall you can use the `goodenuffR` functions to create specific files in your project: 

```{r goodenuffR-funs}
goodenuffR::goodenuff_code() # create code/ folder and files
goodenuffR::goodenuff_data() # create data/ folder and data/raw folder
goodenuffR::goodenuff_doc() # create documentation files
goodenuffR::goodenuff_results() # create results folder 
goodenuffR::goodenuff_collab() # create CITATION, requirements, etc.
```

Let's get started! 

# Packages 

The packages below have been covered in the course slides and exercises. You're not required to know all the details of every package (we can look that up), but you *do* need to know how to install a package. 

```{r packages, message=FALSE, warning=FALSE, eval=TRUE}
library(tidyverse) # for dplyr, tidyr, ggplot2 and others
library(janitor) # for names 
library(skimr) # basic data summaries
```


# Import 

Use the code chunk below to import your data into RStudio. You can replace `Your Raw Data` with your *actual* dataset name, 

```{r import-data, eval=FALSE}
# this will import data in a .csv file, other options include readxl and 
# googlesheets4, and vroom
# Your Raw Data <- readr::read_csv(file = "______________")
```

# Dimensions and structure 

We will start by asking some general questions about the dataset structure and shape. 

## How many columns?

Get the number of columns:

```{r ncol, eval=FALSE}
# ncol(Your Raw Data)
```

```{r length, eval=FALSE}
# length(Your Raw Data)
```

## How many rows? 

Get the number of rows:

```{r nrow, eval=FALSE}
# nrow(Your Raw Data)
```

```{r dim, eval=FALSE}
# dim(Your Raw Data)
```

## Variable names? 

What are the names of the variables/columns in your dataset?

```{r names, eval=FALSE}
# names(Your Raw Data)
```

```{r colnames, eval=FALSE}
# colnames(Your Raw Data)
```


## Variable formats?

How is each column formatted?

```{r str, eval=FALSE}
# str(Your Raw Data)
```

```{r glimpse, eval=FALSE}
# glimpse(Your Raw Data)
```

## Export raw data

Now that we know our data imported correctly, we can export an example of these data to the `data/raw/` folder. 

```{r out_path, eval=FALSE}
fs::dir_create("data/raw/")
# out_path <- paste0("data/raw/", noquote(lubridate::today()), "-Your Raw Data.csv")
# write_csv(x = Your Raw Data, file = out_path)
# verify
fs::dir_tree("data/raw/")
```

# Wrangle

Standardize column names, make sure they're meaningful, and check the contents of each variable.

## Column names 

Use `janitor::clean_names()` to make column names easier to read and program with. 

```{r clean_names, eval=FALSE}
# Your Data <- janitor::clean_names(Your Raw Data)
# glimpse(Your Data)
```

Change the name of `Your Raw Data` to `Your Data` because you've altered the contents of the raw data.

## Summary statistics

Calculate the summary statistics with the `skimr::skim()` function.  

```{r skim-PeopleMap, eval=FALSE}
# skimr::skim(Your Data)
```

## Visualizations

Built *at least two visualizations* of the columns in your dataset. 

## Single variable graphs

View the distribution (or 'shape') of a variable, you can use histograms (`geom_histogram()`), density plots (`geom_density()`), frequency polygons (`geom_freqpoly()`), or box-plots `geom_boxplot()`.

## Multiple variable graphs

View multiple variables with box-plots, ridgeline plots (from the `ggridges` package), or faceting (`facet_wrap()`).

# Data dictionary 

To complete this project, knit your `.html` report and complete a data dictionary for the dataset, This should contain information on each variable and their contents. 

For examples, see the example for the `palmerpenguins::penguins` data.
