
## Separate more complicated rows

Let's take another look at `FandangoSample` below. 

```{r glimpse-FandangoSample}
dplyr::glimpse(FandangoSample)
```

### Shorten column names 

The first thing we want to do is reduce the length of the `rottentomatoes_scores` column name. This is a lot to type, and it technically contains two *versions* of scores (the `%` and `score`/`out of 5`). We'll rename this column `all_rt_vals`. 

```{r all_rt_vals}
FandangoSample <- FandangoSample %>% 
  dplyr::rename(all_rt_vals = rottentomatoes_scores)
utils::head(FandangoSample)
```

### The `sep` argument

If we use `separate_rows()` and leave `sep` empty, it will work (because the default value will match all punctuation characters), but the result isn't what we want: 

```{r separate_rows_ex_05, eval=TRUE}
FandangoSample %>% 
  tidyr::separate_rows(all_rt_vals)
```

As we can see, the `sep` pattern treated each unique non-punctuation/space character as a value. This is why the `%` and all the spaces were been removed, and each element has been placed on a new row.

Let's try this again with a `sep` argument. 

```{r separate_rows_ex_06, eval=TRUE}
FandangoSample %>% 
  tidyr::separate_rows(all_rt_vals, sep = "/")
```

Now we have the percentages and scores on separate rows, but these data still require some additional attention: 

  - **First, we need to include an index for the type of score (`rt_key`) that tells us whether the score in `all_rt_vals` is a `Raw user score (%)` or a `Normalized user score (out of 5)`**    
    - *We can create this with `dplyr::mutate()`, `dplyr::case_when()`, and `stringr::str_detect()`* 

```{r rottentomatoes_key}
FandangoSample %>%
  tidyr::separate_rows(all_rt_vals, sep = "/", convert = TRUE) %>%
  dplyr::mutate(rt_key = dplyr::case_when(
    stringr::str_detect(
      string = all_rt_vals, pattern = "%" # match % 
      ) ~ "Raw user score (%)",
    stringr::str_detect(
      string = all_rt_vals, pattern = "out of" # match 'out of'
      ) ~ "Normalized user score (out of 5)",
  ))
```
    
  - **Next we need to add a zero to the front of the percentages in `all_rt_vals`**   
    - *We can do this with another pass to `dplyr::mutate()`, `dplyr::case_when()`, and `stringr::str_detect()`*

```{r}
FandangoSample %>%
  tidyr::separate_rows(all_rt_vals, sep = "/", convert = TRUE) %>%
  dplyr::mutate(rt_key = dplyr::case_when(
    stringr::str_detect(
      string = all_rt_vals, pattern = "%" # match % 
      ) ~ "Raw user score (%)",
    stringr::str_detect(
      string = all_rt_vals, pattern = "out of" # match 'out of'
      ) ~ "Normalized user score (out of 5)",
  ))  %>%
  dplyr::mutate(all_rt_vals = dplyr::case_when(
    stringr::str_detect(
      all_rt_vals, "%") ~ paste0("0.", all_rt_vals),
    TRUE ~ all_rt_vals
  )) 
```

  - **Finally, we need to remove the `%` and ` out of 5` text from the numbers in `all_rt_vals`**  
    - *We can do this with `dplyr::mutate()` and `stringr::str_remove_all()`*  

```{r}
FandangoSample %>%
  tidyr::separate_rows(all_rt_vals, sep = "/", convert = TRUE) %>%
  dplyr::mutate(rt_key = dplyr::case_when(
    stringr::str_detect(all_rt_vals, "%") ~ "Raw user score (%)",
    stringr::str_detect(all_rt_vals, "out of") ~ "Normalized user score (out of 5)",
  ))  %>%
  dplyr::mutate(all_rt_vals = dplyr::case_when(
    stringr::str_detect(all_rt_vals, "%") ~ paste0("0.", all_rt_vals),
    TRUE ~ all_rt_vals
  )) %>%
  dplyr::mutate(
    all_rt_vals = stringr::str_remove_all(all_rt_vals, "%| out of 5"),
    all_rt_vals = as.double(all_rt_vals)
  )
```

Now we have 

***
