---
title: "Reshaping and Joining Data in R"
comment: "*transpose, pivot, split, and combine data*"
output: 
  html_document: 
    toc: yes
    toc_float: yes
    toc_depth: 6
    number_sections: yes
    code_folding: show
    theme: yeti
    df_print: paged
    
always_allow_html: true
---

```{r , include=FALSE}
library(tidyverse)
library(skimr)
library(fivethirtyeight)
library(fivethirtyeightdata)
# create data folder
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE,
                      tidy = FALSE, 
                      fig.path = "img/")
```

# Objectives 

This document outlines some common data reshaping tasks with `tidyr` and introduces the concept of *tidy data*. 

# Materials 

View the slides for this section [here](https://mjfrigaard.github.io/csuc-data-journalism/slides.html).

Read more about `dplyr` [here](https://dplyr.tidyverse.org/index.html) and [`tidyr`](https://tidyr.tidyverse.org/) on the tidyverse website, or in the [Data Transformation](https://r4ds.had.co.nz/transform.html) chapter of R for Data Science. 

## Load packages 

`tidyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. 

```{r packages, eval=FALSE}
install.packages("tidyverse")
library(tidyverse)
library(fivethirtyeight)
```


## Tidy data {.tabset}

There are two principles for tidy data: 

1. Columns hold *variables*    
2. Rows hold *observations*    
3. Cells (the intersection of rows and columns) hold *values*  

For these first few exercises, we're going to be using data from the [`fivethirtyeight` package](https://fivethirtyeight-r.netlify.app/).


### `EndorseSample`

There are multiple ways data can be 'less-than tidy.' We'll start with a common example, which is two variables (and their values) stored in a single column.  

Below is a subset of the `endorsements_2020` dataset named `EndorseSample`. Read more about the `endorsements_2020` dataset by typing `??endorsements_2020` into the console. 

```{r EndorseSample, message=FALSE, warning=FALSE}
EndorseSample <- readr::read_csv("https://bit.ly/3x6wfTg")
head(EndorseSample)
```

Notice the `location` variable has two values (`state` and `city`) stored in the same column. 

### `FandangoSample`

In another example of un-tidy data, we'll look at the dataset below which comes from the `fandango` dataset. 

```{r FandangoSample, message=FALSE, warning=FALSE}
FandangoSample <- readr::read_csv(file = "https://bit.ly/3kS34yj")
head(FandangoSample)
```

This dataset has two *similar* scores combined into a single column `rottentomatoes_scores`

+ `rottentomatoes_user`: The Rotten Tomatoes user score for the film (expressed as a percent)

+ `rt_user_norm`: The Rotten Tomatoes user score for the film , normalized to a 0 to 5 point system (expressed as 'score `out of 5`')

### `DailyShowSample`

Below is a sample from the `daily_show_guests` dataset. The guests names are separated into `first_name` and `last_name` columns. 

```{r DailyShowSample, message=FALSE, warning=FALSE}
DailyShowSample <- readr::read_csv("https://bit.ly/3HGx8XG")
head(DailyShowSample)
```

***

# Reshaping Columns and Rows 

In the next few exercises, we'll go over how (and *when*) to use `separate()`, `separate_rows()`, and `unite()` these variables.

## `separate()`  {.tabset}

This allows us to split columns apart into multiple variables. 

Take a look at `EndorseSample` below:

```{r glimpse-EndorseSample, eval=TRUE}
EndorseSample
```


### exercise 

Separate `location` into `state` and `city`

```{r separate-ex, eval=FALSE}
EndorseSample %>% 
    separate(col = ___, into = c("_______", "_______"), sep = " ")
```


### solution 

```{r separate-sol}
EndorseSample %>% 
    separate(col = location, into = c("state", "city"), sep = ", ")
```

***

## `separate_rows()`  {.tabset}

`separate_rows()` is similar to `separate()`, but drops the values down the rows. This is commonly used when we have more than one *similar* value in each column. 

Take a look at `FandangoSample` below. 

```{r glimpse-FandangoSample}
FandangoSample
```

### exercise 

Put all the `value`s into their own cell.

```{r separate_rows-ex, eval=FALSE}
FandangoSample %>% 
  separate_rows(____________________, sep = ___)
```


### solution

This isn't as clear as it could be, so we will use `mutate()`, `case_when()`, and `str_detect()` to create an index variable for the `rottentomatoes_scores`, which we will call `rottentomatoes_score`. We'll also remove the `%` and `out of 5` text from `rottentomatoes_scores`, and rename this to `rottentomatoes_value`. 

```{r FandangoSampleTidy}
FandangoSampleTidy <- FandangoSample %>% 
  separate_rows(rottentomatoes_scores, sep = "/", convert = TRUE) %>% 
  mutate(rottentomatoes_score = case_when(
    str_detect(rottentomatoes_scores, "%") ~ "raw user score (%)", 
    str_detect(rottentomatoes_scores, "out of 5") ~ "normalized user score (out of 5)", 
    TRUE ~ NA_character_),
    rottentomatoes_value = str_remove_all(string = rottentomatoes_scores, "%|out of 5"), 
    rottentomatoes_value = as.numeric(rottentomatoes_value)
  ) %>% 
  select(-rottentomatoes_scores)
FandangoSampleTidy
```

Notice this new dataset, `FandangoSampleTidy`, has twice the number of rows as the original `FandangoSample`. 

```{r review-FandangoSample}
FandangoSample
```

***

## `unite()` {.tabset} 

Just like `separate`/`separate_rows` allows us to split up columns, `unite()` lets us stick them back together! Take a look at `DailyShowSample`.

```{r review-DailyShowSample}
DailyShowSample
```

We can see the guests have had their names in separate columns (`first_name` and `last_name`).


### exercise 

Unite `first_name` and `last_name` into `guest_name` (don't remove the previous columns with `remove = FALSE`).

```{r unite, eval=FALSE}
DailyShowSample %>% 
    unite(col = "_______", first_name:last_name, sep = " ", remove = _____)
```

### solution

Unite `first_name` and `last_name` into `guest_name` (don't remove the previous columns with `remove = FALSE`).

```{r unite-solution-3}
DailyShowSample %>% 
    unite(col = "guest_name", first_name:last_name, sep = " ", remove = FALSE)
```

***

# Joins

Joins give us the ability to combine multiple datasets on a common column. We're going to be using the [`starwarsdb` package](https://github.com/gadenbuie/starwarsdb), which contains data from the [Star Wars API](https://swapi.dev/). 

We'll load the data below: 

```{r starwarsdb-package, message=FALSE}
library(starwarsdb)
library(dm)
```

## `starwarsdb` {.tabset}

The code below creates and displays the data model for the tables in this package 

```{r StarWarsDataModel, message=FALSE, warning=FALSE}
StarWarsDataModel <- dm_draw(dm = starwars_dm(), 
                             graph_name = "StarWarsDataModel")
StarWarsDataModel
```



### exercise

Load the `starwarsdb::films` dataset, removing the `opening_crawl` column, and filter to only episodes 4-6. Store these data in `StarWarsFilms`

```{r films-ex, eval=FALSE}
StarWarsFilms <- starwarsdb::films %>% 
                    select(-____________) %>% 
                    filter(episode_id >= ___ & episode_id <= ___)
StarWarsFilms
```

### solution 

```{r films-sol}
StarWarsFilms <- starwarsdb::films %>% 
                    select(-c(opening_crawl, director, producer)) %>% 
                    filter(episode_id >= 4 & episode_id <= 6)
StarWarsFilms
```


## `left_join` {.tabset}

A `left_join` keeps all of `x`, and joins it to all matching rows from dataset `y`

### exercise

Left-join `StarWarsFilms` to `starwarsdb::films_vehicles` on `"title"`, then `select()` the `vehicle` column, then `everything()` else. 

Assign to `StarWarsFilmVeh`

```{r StarWarsFilmsPeople-ex, eval=FALSE}
left_join(x = _____________, 
          y = starwarsdb::films_vehicles, 
          by = ______) %>% 
    select(_____________, 
           everything()) -> StarWarsFilmVeh
StarWarsFilmVeh
```

### solution



```{r StarWarsFilmsPeople}
left_join(x = StarWarsFilms, 
          y = starwarsdb::films_vehicles, 
          by = "title") %>% 
  select(vehicle, 
         everything()) -> StarWarsFilmVeh
StarWarsFilmVeh
```

## `inner_join` {.tabset}

The `inner_join()` joins dataset `x` and `y`, and keeps only matching rows from both.

### exercise 

Use an `inner_join` to join `StarWarsFilmVeh` to `starwarsdb::pilots` on `"vehicle"` and assign it to `StarWarsFilmVehPil`.

```{r StarWarsFilmVenPlt-ex, eval=FALSE}
inner_join(x = StarWarsFilmVeh, 
           y = _________, 
           by = _________) -> _____________
StarWarsFilmVehPil
```

### solution



```{r StarWarsFilmVehPil-sol}
inner_join(x = StarWarsFilmVeh, 
           y = starwarsdb::pilots, 
           by = "vehicle") -> StarWarsFilmVehPil
StarWarsFilmVehPil
```

## `right_join` {.tabset}

The `right_join()` includes all the rows in dataset `y`, and joins it the matching rows in dataset `x`.

### exercise

Use a `right_join()` to join `StarWarsFilmVehPlt` to `starwarsdb::films_planets` on `"title"`, keeping all the rows in `starwarsdb::films_planets`. Use `select()` to rename `planet` to `film_planet` and assign this to `StarWarsFilmVehPltPlnt` 

```{r right_join_ex, eval=FALSE}
right_join(x = __________________, 
             y = __________________, 
             by = "title") %>% 
  select(vehicle:release_date,
         film_planet = _______,
         pilot) -> StarWarsFilmVehPltPlnt
```

### solution



```{r right_join-sol}
right_join(x = StarWarsFilmVehPil, 
           y = starwarsdb::films_planets, 
           by = "title") %>% 
  select(vehicle:release_date,
         film_planet = planet, 
         pilot) -> StarWarsFilmVehPltPlnt
StarWarsFilmVehPltPlnt
```

## `anti_join` {.tabset}

The `anti_join()` keeps all rows in dataset `x` that **do not** have a match in dataset `y`. Note the last join created empty values for `vehicle`, `episode_id`, `release_date`, and `pilot`. 

```{r filter-is.na}
StarWarsFilmVehPltPlnt %>% 
  filter(is.na(episode_id))
```

### exercise 

Use the `anti_join()` to join `StarWarsFilmVehPltPlnt` to `StarWarsFilms` and create a dataset with only `title` and `film_planet` (*these should only contain the `title`s not in `StarWarsFilms`*).

```{r anti_join-ex, eval=FALSE}
anti_join(x = StarWarsFilmVehPltPlnt, 
            y = ____________, 
            by = c("title", "episode_id", "release_date")) %>% 
  select(________, ________)
```

### solution 

```{r anti_join-sol}
anti_join(x = StarWarsFilmVehPltPlnt, 
            y = StarWarsFilms, 
            by = c("title", "episode_id", "release_date")) %>% 
  select(title, film_planet)
```



***

# Nested data

Nested data refers to a dataset in which a column is holding a list of data frames (or `tibble`s). These can be useful for manipulating data, and we're going to give an example of using nested data to reshape a column combining two sets of dates.

## Anti-Doping Sanctions Database {.tabset}

`unnest()` deals with nested data, which can be complicated to explain, so we'll use an example. Run the code below to download data from the [Anti-Doping Agency Sanction database](https://www.usada.org/news/sanctions/). 

```{r 01-create-usada-data, message=FALSE, warning=FALSE}
source("code/create-usada-data.R")
```

These data have two dates (`original` and `updated`) in a single column (`sanction_announced`). **We only want `updated` dates in their own columns.**

```{r UsadaBadDates}
UsadaBadDates
```

### `str_split` 

Use the `stringr` package's `str_split` function to split the bad dates on the `"updated"` pattern. 

```{r str_split}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```

This creates a nested column (`sanction_dates`), which we can deal with by `unnest()`ing it. Let's take a better look with `listviewer` (but only look at `athlete` and `sanction_dates`).

```{r listviewer}
library(listviewer)
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_dates) %>% 
    listviewer::jsonedit(mode = "view")
```


### `unnest`

After unnesting, we can see the dates have been repeated down the rows (like `separate_rows()`).

```{r unnest}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    # check this to see what is happening 
        dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `filter` 

We can see all of the updated dates are on a new line, and the original dates still contain the text. Now I can filter out the rows with an original date (because I am only interested in the updated dates).

```{r filter}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # check this to see what is happening 
        dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `str_remove_all`

Now I can clean up the updated `sanction_dates` and format it. 

```{r str_remove_all}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>%
    # check this to see what is happening 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```


### `ggplot2`

***How far can this pipe go??***

Well, when the data are arranged in the way we want, we can test it with a visualization to check the number of sanctions per sport with bad dates. 

```{r visualize, out.height='100%', out.width='100%', fig.height=5, fig.width=7}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
                        
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
  
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
  
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>% 
  
    # count by sport
    dplyr::count(sport, sort = TRUE) %>% 
  
    # rename n 
    rename(`Sport Count` = n,
            Sport = sport) %>% 
    ggplot2::ggplot(aes(x = `Sport Count`, 
                        y = forcats::fct_reorder(as.factor(Sport), `Sport Count`), 
                        fill = as.character(`Sport Count`))) + 
    geom_col(show.legend = FALSE) + 
    labs(x = "Number of Sanctions", y = "Sport",
         title = "Number of Sanctions with Bad Dates",
         caption = "source: https://www.usada.org/news/sanctions/") + 
    ggthemes::theme_fivethirtyeight() + 
    ggthemes::scale_fill_few()
```


