---
title: "Joining Data in R"
description: |
  "*Using `dplyr` to join data tables*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 2
    
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(inspectdf)
library(starwarsdb)
library(dm)
# create data folder
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5.5,
  fig.retina = 2
  )
fs::dir_create("data")
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/reshape-join/')
# tibble printing
options(
  pillar.print_max = 10, 
  pillar.print_min = 10)
```

# Objectives 

This document outlines some common data joining tasks with the `dplyr` package. **If you've ever Googled 'how to perform [x] in R', you've probably discovered there are multiple ways to accomplish [x].** 

The methods I present here are not exhaustive, and I've tried to balance the trade-off between 1) solutions/functions that solve problems with the fewest number of keystrokes, and 2) code that's readable, explicit, and easy to follow.


## Load packages 

`dplyr` is part of the core `tidyverse` packages, so we install and load this meta-package below. We'll also be installing the `inspectdf`, `starwarsdb`, and `dm` packages. 

```{r packages, eval=FALSE}
# install.packages(c("tidyverse", "inspectdf", "starwarsdb", 
#                   "dm", "patchwork", "kableExtra"))
# devtools::install_github("nsgrantham/ggdark")
library(tidyverse)
library(inspectdf)
library(starwarsdb)
library(dm)
library(patchwork)
library(kableExtra)
library(ggdark)
```

## Namespacing

I'll be using the namespace syntax (`package::function()`) just to make it explicitly clear which function belongs to which package, but this isn't necessary if you've run the code block above. 

## Tidy data 

There are three principles for tidy data: 

1. Columns hold *variables*    
2. Rows hold *observations*    
3. Cells (the intersection of rows and columns) hold *values* 


# Joins

Joins give us the ability to combine and filter multiple datasets on common column(s). 

