---
title: "Wrangling USADA data"
description: |
  "*Advanced example of data reshaping and restructuring*"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 3
    
always_allow_html: true
---

```{r , include=FALSE}
library(tidyverse)
library(skimr)
library(fivethirtyeight)
library(fivethirtyeightdata)
# create data folder
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  fig.path = "img/",
  dpi = 320,
  fig.align = "center",
  fig.width = 8.5,
  fig.height = 6,
  fig.retina = 2
  )
fs::dir_create("data")
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/reshape-join/')
```


# Nested data

Nested data refers to a dataset in which a column is holding a list of data frames (or `tibble`s). These can be useful for manipulating data, and we're going to give an example of using nested data to reshape a column combining two sets of dates.

## Anti-Doping Sanctions Database

`unnest()` deals with nested data, which can be complicated to explain, so we'll use an example. 

Below I've created a function (`scrape_usada()`) that scrapes the data from the [Anti-Doping Agency Sanction database](https://www.usada.org/news/sanctions/). 

```{r scrape_usada, echo=FALSE, eval=TRUE}
scrape_usada <- function() {
  usada_url <- "https://www.usada.org/testing/results/sanctions/"
  usada_extraction <- rvest::read_html(usada_url) |> rvest::html_nodes("table")
  usada_raw <- rvest::html_table(usada_extraction[[1]])
  usada <- janitor::clean_names(usada_raw)
  usada <- tibble::add_column(usada, timestamp = Sys.Date())
  return(usada)
}
```

```{r usada_raw}
usada_raw <- scrape_usada()
glimpse(usada_raw)
```

These data have two dates (`original` and `updated`) in a single column (`sanction_announced`). **We only want `updated` dates in their own columns.**

```{r UsadaBadDates}
UsadaBadDates
```

### `str_split` 

Use the `stringr` package's `str_split` function to split the bad dates on the `"updated"` pattern. 

```{r str_split}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```

This creates a nested column (`sanction_dates`), which we can deal with by `unnest()`ing it. Let's take a better look with `listviewer` (but only look at `athlete` and `sanction_dates`).

```{r listviewer}
library(listviewer)
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # only return what we're interested in 
    dplyr::select(athlete, sanction_dates) %>% 
    listviewer::jsonedit(mode = "view")
```


### `unnest`

After unnesting, we can see the dates have been repeated down the rows (like `separate_rows()`).

```{r unnest}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    # check this to see what is happening 
        dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `filter` 

We can see all of the updated dates are on a new line, and the original dates still contain the text. Now I can filter out the rows with an original date (because I am only interested in the updated dates).

```{r filter}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # check this to see what is happening 
        dplyr::select(athlete, sanction_announced, sanction_dates)
```

### `str_remove_all`

Now I can clean up the updated `sanction_dates` and format it. 

```{r str_remove_all}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
    
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>%
    # check this to see what is happening 
    dplyr::select(athlete, sanction_announced, sanction_dates)
```


### `ggplot2`

***How far can this pipe go??***

Well, when the data are arranged in the way we want, we can test it with a visualization to check the number of sanctions per sport with bad dates. 

```{r visualize, out.height='100%', out.width='100%', fig.height=5, fig.width=7}
UsadaBadDates %>%
        dplyr::mutate(sanction_dates = 
                        
    # 1) split this on the "updated" pattern
             stringr::str_split(string = sanction_announced, 
                                pattern = "updated")) %>% 
  
    # 2) convert the output from split into multiple rows
        tidyr::unnest(sanction_dates) %>% 
  
    # 3) remove the "original" dates 
        dplyr::filter(!str_detect(string = sanction_dates, 
                                  pattern = "original")) %>% 
    # 4) remove the colon from sanction_dates
        dplyr::mutate(sanction_dates = stringr::str_remove_all(
                                                      string = sanction_dates,
                                                      pattern = ":"),
                      
                      # 5) remove any whitespace
                      sanction_dates = stringr::str_trim(sanction_dates),
                      
                      # 6) format as date
                      sanction_dates = lubridate::mdy(sanction_dates)) %>% 
  
    # count by sport
    dplyr::count(sport, sort = TRUE) %>% 
  
    # rename n 
    rename(`Sport Count` = n,
            Sport = sport) %>% 
    ggplot2::ggplot(aes(x = `Sport Count`, 
                        y = forcats::fct_reorder(as.factor(Sport), `Sport Count`), 
                        fill = as.character(`Sport Count`))) + 
    geom_col(show.legend = FALSE) + 
    labs(x = "Number of Sanctions", y = "Sport",
         title = "Number of Sanctions with Bad Dates",
         caption = "source: https://www.usada.org/news/sanctions/") + 
    ggthemes::theme_fivethirtyeight() + 
    ggthemes::scale_fill_few()
```


