---
title: "Fixing bad graphs"
description: |
  "From Medium artile, '[Mistakes, we’ve drawn a few'](https://medium.economist.com/mistakes-weve-drawn-a-few-8cdd8a42d368)'"
date: "`r Sys.Date()`"
output:
  distill::distill_article:
    toc: true
    toc_float: yes
    toc_depth: 4

always_allow_html: true
    
---


```{r setup, include=FALSE}
library(fs)
library(tidyr)
library(dplyr)
library(ggplot2)
library(tidytuesdayR)
library(zipcodeR)
library(tigris)
library(readr)
library(stringr)
library(lubridate)
library(patchwork)
library(grid)
## Global options
knitr::opts_chunk$set(
  cache = FALSE,
  echo = TRUE,
  collapse = FALSE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE,
  dev = "ragg_png",
  out.width = '85%',
  dpi = 320,
  fig.align = "center",
  fig.width = 7.5,
  fig.height = 5,
  fig.retina = 2
  )
fs::dir_create("data")
knitr::opts_knit$set(root.dir = '/Users/mjfrigaard/Dropbox/@projects/high-priority/@data-journalism/course-website/FA22-JOUR301/drafts/fixing-bad-graphs/')
```

## Install packages 

Put the code from `00-inst.R` below:

```{r 00-inst, message=FALSE, warning=FALSE}
# install.packages(c("tidyverse", "janitor", "skimr", "fs", "ggthemes"))
library(tidyverse)
library(janitor)
library(skimr)
library(fs)
library(ggthemes)
```

## Import data 

Put the code from `01.import.R` below: 

```{r 01-import}
# create data folder
fs::dir_create("data/")
# create 'data/raw/' folder
fs::dir_create("data/raw/")

# data urls
balance_url <- "https://bit.ly/3hRzrKS"
brexit_url <- "https://bit.ly/3s2wqMx"
corbyn_url <- "https://bit.ly/35mgYRB"
pensions_url <- "https://bit.ly/2MNAvEp"

# data destination file paths
balance_destfile <- "data/raw/Balance.csv"
brexit_destfile <- "data/raw/Brexit.csv"
corbyn_destfile <- "data/raw/Corbyn.csv"
pensions_destfile <- "data/raw/Pensions.csv"

# download the files using the url and destfile
download.file(url = balance_url, destfile = balance_destfile)
download.file(url = brexit_url, destfile = brexit_destfile)
download.file(url = corbyn_url, destfile = corbyn_destfile)
download.file(url = pensions_url, destfile = pensions_destfile)

# import data
Balance <- readr::read_csv(balance_destfile)
Brexit <- readr::read_csv(brexit_destfile)
Corbyn <- readr::read_csv(corbyn_destfile)
Pensions <- readr::read_csv(pensions_destfile)
```

## Plot a histogram of average facebook likes with `qplot()`

Use the `Corbyn` dataset to view the `avg_facebook_likes` variable.

```{r avg_facebook_likes}
Corbyn %>%
  ggplot2::qplot(x = avg_facebook_likes, data = .)
```

## Plot a box-plot of `pop_65_percent`

Use `qplot()` to graph the `pop_65_percent` variable from the `Pensions` dataset: 

```{r pop_65_percent}
Pensions %>%
  # the variable
  ggplot2::qplot(
    x = pop_65_percent,
    y = " ",
    data = .,
    geom = "boxplot"
  )
```

## Create a box-plot of `value` and `country` 

Use the Pensions dataset to plot `value` (on the x) and `country` (on the `y`)

```{r value-country}
Balance %>%
  ggplot2::qplot(
    x = value,
    y = country,
    data = .,
    geom = "boxplot"
  )
```

## Wrangle `date` variable in Brexit 

Format the `date` variable in the `Brexit` dataset with `lubridate::dmy()`

```{r lubridate-date}
Brexit <- Brexit %>% mutate(date = lubridate::dmy(date))
glimpse(Brexit)
```


## Create a scatter-plot of `date` and `percent_responding_right`

Use the `Brexit` dataset to plot date (on the x) and `percent_responding_right` (on the `y`)

```{r date-percent_responding_right}
Brexit %>%
  qplot(
    x = date,
    y = percent_responding_right, data = .
  )
```

## Build plots layer-by-layer with `ggplot()`

Run the code below to see how `tidyr::pivot_longer()` works.

```{r}
Brexit
```


```{r pivot_longer}
Brexit %>%
  tidyr::pivot_longer(
    cols = -date,
    names_to = "poll",
    values_to = "percent"
  )
```

Now pass the pivoted data to `qplot()` and view the result. 

```{r pivot_longer-qplot}
Brexit %>%
  pivot_longer(
    cols = -date,
    names_to = "poll",
    values_to = "percent"
  ) %>%
  ggplot2::qplot(
    x = date,
    y = percent,
    group = poll,
    data = .,
    geom = "line",
    color = poll
  )
```


Wrangle the `Brexit` data below: 

Assign the `tidyr::pivot_longer()` output to `TidyBrexit`

```{r TidyBrexit}
TidyBrexit <- Brexit %>%
  pivot_longer(-date,
    names_to = "poll",
    values_to = "percent"
  )
head(TidyBrexit)
```

## Map variables to aesthetics 

Use the `TidyBrexit` `data.frame` to build the `ggp_brexit` plot object layer. 

```{r ggp_brexit}
ggp_brexit <- TidyBrexit %>%
  ggplot(mapping = aes(x = date, y = percent, group = poll))
ggp_brexit
```

## Add the geom

Add the `ggplot2::geom_smooth()` to `ggp_brexit`

```{r geom_smooth}
ggp_brexit + 
  ggplot2::geom_smooth()
```

## Add the color aethetic to `poll`

```{r color-poll}
ggp_brexit + 
  ggplot2::geom_smooth(aes(color = poll))
```

Add the `se = FALSE` and `show.legend = FALSE` and create the `ggp_brexit_smooth` object. 

```{r ggp_brexit_smooth}
ggp_brexit_smooth <- ggp_brexit +
  geom_smooth(aes(color = poll), 
    se = FALSE, show.legend = FALSE)
ggp_brexit_smooth
```

## Add the points with `ggplot2::geom_point()`

Add the points with `ggplot2::geom_smooth()`, but include the `show.legend = FALSE` and `alpha = 1/3`

```{r geom_point}
ggp_brexit_smooth +
  geom_point(aes(color = poll), 
    show.legend = FALSE, alpha = 1 / 3)
```

## Set the aesthetics globally 

Rebuild the plot with the `TidyBrexit` dataset, mapping `date`, `percent`, and `poll` globally. 

+ add the `ggplot2::geom_smooth()` layer with the `se = FALSE` and `show.legend = FALSE`   

+ add the `ggplot2::geom_point()` layer with `alpha = 1/3` and `show.legend = FALSE`   

```{r ggp_brexit_global}
ggp_brexit_global <- TidyBrexit %>%
  ggplot(mapping = aes(x = date, y = percent, group = poll)) +
  geom_smooth(aes(color = poll), se = FALSE, show.legend = FALSE) +
  geom_point(aes(color = poll), alpha = 1 / 3, show.legend = FALSE)
ggp_brexit_global
```


## Add colors 

Add the colors `cornflowerblue` and `firebrick3` to the `ggp_brexit_global` layer and create a new layer called `ggp_brexit_global_colors`. 

```{r ggp_brexit_global_colors}
ggp_brexit_global_colors <- ggp_brexit_global +
  scale_color_manual(values = c("cornflowerblue", "firebrick3"))
ggp_brexit_global_colors
```

## Add text 

Use the `ggplot2::geom_text()` layer to add the text to the `ggp_brexit_global_colors` layer, and name the new plot `ggp_brexit_global_colors_text`


```{r ggp_brexit_global_colors_text}
ggp_brexit_global_colors_text <- ggp_brexit_global_colors +
  annotate(geom = "text",
    label = "Wrong", 
    color = "firebrick3",
    x = as.Date("2018-01-01"), 
    y = 45.5
  ) +
  annotate(geom = "text",
    label = "Right", 
    color = "cornflowerblue",
    x = as.Date("2018-01-01"),
    y = 42.5
  )
ggp_brexit_global_colors_text
```

## Move the `y` axis 

Move the y axis to the right side of the graph with `ggplot2::scale_y_continuous()`

```{r ggp_brexit_global_colors_text_scale_y}
ggp_brexit_global_colors_text_scale_y <- ggp_brexit_global_colors_text +
  ggplot2::scale_y_continuous(position = "right")
ggp_brexit_global_colors_text_scale_y
```

## Add labels 

Use the `ggplot2::labs()` function to create the labels layer. 

```{r labs_eco}
labs_eco <- ggplot2::labs(
  title = "Bremorse",
  subtitle = "'In hindsight, do you think Britain was right or wrong to vote to leave the EU?'",
  caption = "Source: NatCen Social Research",
  x = " ",
  y = "Percent (%)"
)
```

Add the labels layer to `ggp_brexit_global_colors_text_scale_y` and create a new plot titled, `ggp_brexit_labs`. 

```{r ggp_brexit_labs}
ggp_brexit_labs <- ggp_brexit_global_colors_text_scale_y + labs_eco
ggp_brexit_labs
```


## Add theme 

```{r install-ggthemes}
# install.packages("ggthemes")
library(ggthemes)
```

Add the `ggthemes::theme_economist_white()` layer and set the following arguments: 

- `gray_bg = FALSE`  
- `base_size = 12`  
- `base_family = "Verdana"`  

```{r theme_economist_white}
ggp_brexit_labs +
  ggthemes::theme_economist_white(
    gray_bg = FALSE,
    base_size = 12,
    base_family = "Verdana"
  )
```
